---
title: "02 — Data Preparation: Tick Quotes to Event-Time Series"
subtitle: "ACD-GARCH Intraday Volatility & Market Risk Forecasting — BP LSE Tick Data"
author: "Sourodeep Roy"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

## Purpose

This notebook ingests raw BP LSE tick quote data (April–June 2020) and
produces clean event-time series ready for ACD modelling.

**Key design decision:** Events are defined as *midquote changes* — quote
updates where the mid-price `(ask + bid) / 2` moves. This filters out
volume-only updates that carry no price information and would produce
spurious zero-return durations in the ACD model.

- **Inputs:**  `data/raw/BPGBGBX_Ticks_2020.04.01_2020.06.30.csv`
- **Outputs:**
  - `data/processed/tick_clean.rds`   — full filtered quote data (all months)
  - `data/processed/events_apr.rds`   — April event-time series (training)
  - `data/processed/events_may.rds`   — May event-time series (test)
  - `data/processed/events_jun.rds`   — June event-time series (test)

---

## 1. Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5)
```

```{r load-packages}
# Source the project setup so all packages and global options are loaded.
# This ensures timezone, seed, and package state are consistent with 01_setup.Rmd.
suppressPackageStartupMessages({
  library(data.table)
  library(lubridate)
  library(ggplot2)
  library(here)
})
```

---

## 2. Load Raw Data

```{r load-data}
# Construct path relative to project root using here::here().
# This ensures the notebook runs identically on any machine.
data_path <- here::here("data", "raw", "BPGBGBX_Ticks_2020.04.01_2020.06.30.csv")

stopifnot("Raw data file not found — check data/raw/" = file.exists(data_path))

# fread() is used for speed on 1M+ row files.
# data.table is the right tool here — base read.csv would be ~10x slower.
tick_raw <- fread(data_path)

cat("Raw rows loaded:", format(nrow(tick_raw), big.mark = ","), "\n")
cat("Columns:", paste(names(tick_raw), collapse = ", "), "\n")
```

```{r rename-columns}
# Rename to clean, lowercase, no-space column names for easier referencing
setnames(tick_raw,
  old = c("Time (Europe/London)", "Ask", "Bid", "AskVolume", "BidVolume"),
  new = c("time_raw", "ask", "bid", "ask_volume", "bid_volume")
)
```

---

## 3. Parse Timestamps

```{r parse-timestamps}
# Timestamps arrive as character strings with millisecond precision,
# e.g. "2020.04.01 08:00:11.526". We parse these to POSIXct so we can
# compute inter-event durations accurately.
#
# Format: %Y.%m.%d %H:%M:%OS — the %OS handles fractional seconds.
# tz = "Europe/London" is critical: the period spans the BST transition
# (clocks go forward on 29 March 2020, before our sample; we are already in BST
# for the full April-June window, but we keep the tz explicit throughout).

tick_raw[, timestamp := as.POSIXct(
  trimws(as.character(time_raw)),
  format = "%Y.%m.%d %H:%M:%OS",
  tz     = "Europe/London"
)]

# Hard stop if any timestamps failed to parse — silent NA timestamps would
# produce incorrect durations without any downstream error message
n_failed <- sum(is.na(tick_raw$timestamp))
if (n_failed > 0) stop(n_failed, " timestamps failed to parse. Check raw file format.")

# Drop the original raw time column — we no longer need it
tick_raw[, time_raw := NULL]

# Show ms precision is preserved (display-only — does not affect calculations)
options(digits.secs = 3)
head(tick_raw[, .(
  timestamp_display = format(timestamp, "%Y-%m-%d %H:%M:%OS3"),
  ask, bid
)], 6)
```

---

## 4. Sort and Time Integrity Checks

```{r sort-and-check}
# ACD models require a strictly ordered time series.
# We sort by timestamp and then check for any time reversals.
setorder(tick_raw, timestamp)

# Check 1: time must never go backwards
n_reversals <- sum(diff(as.numeric(tick_raw$timestamp)) < 0)
stopifnot("Time reversals detected — check raw data ordering" = n_reversals == 0)
cat("Time reversals:", n_reversals, "(none expected)\n")

# Check 2: duplicate timestamps — report count but do NOT hard-stop.
# Multiple quote updates within the same millisecond are normal in
# high-frequency data (e.g. simultaneous bid and ask updates).
# These are handled naturally when we filter to midquote-change events.
n_dup <- sum(duplicated(tick_raw$timestamp))
cat("Duplicate timestamps:", format(n_dup, big.mark = ","),
    "(expected — multiple updates can share a millisecond)\n")
```

---

## 5. Quote Sanity Checks

```{r quote-checks}
# Prices must be positive and ask >= bid (no crossed quotes)
stopifnot(
  "Negative ask prices found"  = all(tick_raw$ask  > 0),
  "Negative bid prices found"  = all(tick_raw$bid  > 0),
  "Crossed quotes found (ask < bid)" = all(tick_raw$ask >= tick_raw$bid)
)

# Spread distribution — identifies stale or anomalous quotes
tick_raw[, spread := ask - bid]
cat("\nBid-ask spread summary (pence):\n")
print(summary(tick_raw$spread))

# Zero-spread quotes are unusual in quote data and may indicate stale ticks
cat("Zero-spread quotes:", sum(tick_raw$spread == 0), "\n")
```

---

## 6. Filter to Regular Trading Hours

```{r filter-trading-hours}
# We retain only regular LSE trading days and hours.
# Trading window: 08:00:00 to 16:30:00 London time.
# Weekdays only (Monday = 1, Friday = 5 with week_start = 1).

# UK bank holidays in our sample window (source: Bank of England / LSE calendar)
# These days have no regular trading and must be excluded.
uk_holidays_2020 <- as.Date(c(
  "2020-04-10",  # Good Friday
  "2020-04-13",  # Easter Monday
  "2020-05-08",  # VE Day bank holiday (special 2020)
  "2020-05-25"   # Spring bank holiday
))

# Add date column for filtering
tick_raw[, date := as.Date(timestamp, tz = "Europe/London")]

# Numeric intraday time in seconds from midnight — faster and safer than
# string comparison (avoids format-dependent fragility on large datasets)
tick_raw[, intraday_sec := hour(timestamp) * 3600L +
                           minute(timestamp) * 60L +
                           as.integer(second(timestamp))]

open_sec  <- 8L  * 3600L           # 08:00:00
close_sec <- 16L * 3600L + 30L * 60L  # 16:30:00

tick_clean <- tick_raw[
  wday(timestamp, week_start = 1) <= 5 &     # Monday–Friday only
  !(date %in% uk_holidays_2020)             & # exclude bank holidays
  intraday_sec >= open_sec                  & # at or after 08:00
  intraday_sec <= close_sec                   # at or before 16:30
]

cat("Rows before filter:", format(nrow(tick_raw),   big.mark = ","), "\n")
cat("Rows after filter: ", format(nrow(tick_clean), big.mark = ","), "\n")
cat("Rows removed:      ", format(nrow(tick_raw) - nrow(tick_clean), big.mark = ","), "\n")
```

```{r filter-verification}
# Verify the filter worked correctly
stopifnot(all(wday(tick_clean$timestamp, week_start = 1) <= 5))
stopifnot(!any(tick_clean$date %in% uk_holidays_2020))

time_range <- range(tick_clean$intraday_sec)
cat("Intraday time range:", time_range[1], "to", time_range[2], "seconds\n")
cat("  = ", sprintf("%02d:%02d", time_range[1] %/% 3600, (time_range[1] %% 3600) %/% 60),
    "to", sprintf("%02d:%02d", time_range[2] %/% 3600, (time_range[2] %% 3600) %/% 60), "\n")

cat("Trading days covered:", uniqueN(tick_clean$date), "\n")
cat("Date range:", format(min(tick_clean$date)), "to", format(max(tick_clean$date)), "\n")
```

---

## 7. Intraday Clock

```{r intraday-clock}
# Compute seconds elapsed since 08:00:00 (market open) for each tick.
# We anchor to a FIXED 08:00:00 reference — NOT the first observed tick of the day.
# Using the first tick would give different origins on different days
# (e.g. a quiet day where first tick arrives at 08:00:45).

tick_clean[, trading_seconds := intraday_sec - open_sec]

# Sanity: trading_seconds must be >= 0 and <= 30600 (8.5 hours × 3600)
stopifnot(all(tick_clean$trading_seconds >= 0))
stopifnot(all(tick_clean$trading_seconds <= 30600))
```
```{r sub-second-precision}
head(format(tick_raw$timestamp, "%Y-%m-%d %H:%M:%OS6"), 10)
```

---

## 8. Mid-Price Construction

```{r mid-price}
# We use the quote mid-price as our price proxy.
# Mid-price = (ask + bid) / 2
# This reduces bid-ask bounce compared to using either side alone.
# All subsequent returns and durations are based on this series.

tick_clean[, mid := (ask + bid) / 2]

# Drop the spread column computed earlier (was for diagnostics only)
tick_clean[, spread := NULL]

# Reorder columns for readability
setcolorder(tick_clean, c("timestamp", "date", "trading_seconds",
                          "bid", "ask", "mid",
                          "bid_volume", "ask_volume"))
```

---

## 9. Month Split

### Train / Test Split Design

We use **April 2020 as the training period** and **May–June 2020 as the 
out-of-sample test period**. This design is justified on four grounds:

1. **No look-ahead bias:** A risk manager at end-April would have access 
   only to April data. Our design replicates this exactly.

2. **Economic severity:** April 2020 captures the peak of the COVID oil 
   price crash (Brent crude went negative on April 20). Training on the 
   most volatile month and testing on calmer conditions is a conservative 
   and credible design — the model is not given an easy test.

3. **Statistical sufficiency:** April provides 1,264,699 events across 
   20 trading days — far more than needed to identify ACD(2,2) and 
   GARCH(1,1) parameters reliably.

4. **Literature precedent:** Single fixed-window training with rolling 
   out-of-sample evaluation is standard in the ACD and high-frequency 
   GARCH literature (Engle & Russell 1998; Andersen & Bollerslev 1998).

All model parameters are estimated on April only. No refitting occurs 
on May or June data. ACD recursion and GARCH rolling forecasts for 
May–June use April-estimated parameters exclusively.

```{r month-split}
# Split into training (April) and test (May, June) sets.
# All ACD and GARCH models are fitted on April only.
# May and June are strictly out-of-sample.

tick_clean[, month := format(date, "%Y-%m")]

tick_apr <- tick_clean[month == "2020-04"]
tick_may <- tick_clean[month == "2020-05"]
tick_jun <- tick_clean[month == "2020-06"]

stopifnot(nrow(tick_apr) > 0, nrow(tick_may) > 0, nrow(tick_jun) > 0)

cat(sprintf("%-12s %10s %8s\n", "Month", "Ticks", "Days"))
cat(sprintf("%-12s %10s %8s\n", "April",
            format(nrow(tick_apr), big.mark=","),
            uniqueN(tick_apr$date)))
cat(sprintf("%-12s %10s %8s\n", "May",
            format(nrow(tick_may), big.mark=","),
            uniqueN(tick_may$date)))
cat(sprintf("%-12s %10s %8s\n", "June",
            format(nrow(tick_jun), big.mark=","),
            uniqueN(tick_jun$date)))
```

---

## 10. Event Definition and Duration Construction

```{r make-event-function}
# Events = midquote changes.
# We filter out volume-only quote updates (where ask and bid don't move).
# This is the correct event definition for a quote-based dataset.
#
# For each event we compute:
#   duration  = seconds since the previous event (within the same trading day)
#   ret_mid   = log return of the mid-price since the previous event
#
# The by = date grouping ensures we never compute a duration or return
# that spans overnight — the first event of each day is correctly dropped.

make_event_data <- function(dt) {
  dt <- data.table::copy(dt)
  data.table::setorder(dt, date, timestamp)

  # Keep only rows where the mid-price changed since the previous tick.
  # shift() produces NA for the first row of the full table — this row
  # is naturally excluded by the != comparison (NA != anything = NA = FALSE).
  dt <- dt[mid != data.table::shift(mid)]

  # Inter-event duration in seconds, computed within each trading day.
  # shift() within by = date gives NA for the first event of each day,
  # which we then drop along with any zero-duration ties.
  dt[, duration := as.numeric(
    difftime(timestamp, data.table::shift(timestamp), units = "secs")
  ), by = date]
  dt <- dt[!is.na(duration) & duration > 0]

  # Event-time log return of the mid-price.
  # Computed within each day so overnight gaps don't contaminate returns.
  dt[, ret_mid := c(NA_real_, diff(log(mid))), by = date]
  dt <- dt[!is.na(ret_mid)]

  dt[]
}

evt_apr <- make_event_data(tick_apr)
evt_may <- make_event_data(tick_may)
evt_jun <- make_event_data(tick_jun)

cat(sprintf("%-12s %10s\n", "Month", "Events"))
cat(sprintf("%-12s %10s\n", "April",  format(nrow(evt_apr), big.mark=",")))
cat(sprintf("%-12s %10s\n", "May",    format(nrow(evt_may), big.mark=",")))
cat(sprintf("%-12s %10s\n", "June",   format(nrow(evt_jun), big.mark=",")))
```
# NOTE: ~0.89% of April ticks removed as volume-only updates (symmetric 
# bid-ask moves that leave mid unchanged). Low rate expected — Dukascopy 
# pre-filters feed to bid/ask change events only.
# NOTE: Duration floor at ~50ms likely reflects LSE Minimum Quote Life (MQL)
# requirement. Affects ~25% of durations. Relevant for ACD shape parameter.

```{r event-sanity-checks}
# All durations must be strictly positive and finite
stopifnot(
  "Non-positive durations in April events" = all(evt_apr$duration > 0),
  "Non-finite durations in April events"   = all(is.finite(evt_apr$duration)),
  "Non-finite returns in April events"     = all(is.finite(evt_apr$ret_mid))
)

cat("April duration summary (seconds):\n")
print(summary(evt_apr$duration))

cat("\nApril return summary:\n")
print(summary(evt_apr$ret_mid))
```

---

## 11. Exploratory Analysis

```{r daily-event-counts}
# Events per trading day — helps us spot thin trading days or data gaps
evt_counts_apr <- evt_apr[, .N, by = date][order(date)]

ggplot(evt_counts_apr, aes(x = date, y = N)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  labs(
    title    = "Midquote-change events per trading day — April 2020",
    subtitle = "Each bar = one trading day. Low counts may indicate thin markets or data gaps.",
    x = "Date", y = "Number of events"
  ) +
  theme_minimal(base_size = 12)
```

```{r duration-distribution}
# Duration distribution on log scale — we expect a right-skewed distribution
# consistent with Weibull or Burr innovations (not Exponential).
# The shape here motivates the Weibull ACD specification in notebook 04.

ggplot(evt_apr, aes(x = duration)) +
  geom_histogram(bins = 120, fill = "steelblue", alpha = 0.7, colour = "white") +
  scale_x_log10() +
  labs(
    title    = "Distribution of inter-event durations — April 2020 (log scale)",
    subtitle = "Right skew and heavy tail are expected in high-frequency duration data.",
    x        = "Duration (seconds, log scale)",
    y        = "Count"
  ) +
  theme_minimal(base_size = 12)
```

```{r midquote-path}
# Visual check of the mid-price path on one representative day.
# If timestamps or the intraday clock are wrong, this plot will look broken.

sample_date <- as.Date("2020-04-01")

ggplot(tick_clean[date == sample_date], aes(x = trading_seconds / 3600, y = mid)) +
  geom_line(colour = "steelblue", linewidth = 0.4) +
  scale_x_continuous(
    breaks = 0:8,
    labels = function(x) sprintf("%02d:%02d", 8L + as.integer(x), 0L)
  ) +
  labs(
    title    = paste("BP plc mid-price —", sample_date),
    subtitle = "Source: Dukascopy best bid/ask quotes. Events = midquote changes.",
    x        = "London time",
    y        = "Mid-price (GBX)"
  ) +
  theme_minimal(base_size = 12)
```

```{r duration-tail-quantiles}
# Extreme duration quantiles — these will inform the winsorisation bounds
# we set in notebook 03. The 99.9th and 99.99th percentile values are
# particularly relevant for detecting outlier gaps.
cat("April duration quantiles (seconds):\n")
print(quantile(evt_apr$duration, probs = c(0.25, 0.5, 0.75, 0.95, 0.99, 0.999, 0.9999)))
```

```{r intraday-activity-pattern}
# Average number of events per 5-minute bucket across all April trading days.
# This reveals the intraday seasonality in market activity — a key stylised
# fact that motivates the diurnal adjustment in notebook 03.
# We expect high activity at open (08:00-09:00), declining through midday,
# and a secondary peak near close (15:30-16:30).

# Assign each event to a 5-minute bucket (0 to 101 for 8.5 hours)
evt_apr[, bucket_5m := trading_seconds %/% 300L]

# Average events per bucket across all trading days
activity_profile <- evt_apr[, .N, by = .(date, bucket_5m)][
  , .(avg_events = mean(N)), by = bucket_5m][order(bucket_5m)]

# Convert bucket index back to a readable time label
activity_profile[, time_label := sprintf("%02d:%02d",
  8L + (bucket_5m * 5L) %/% 60L,
  (bucket_5m * 5L) %% 60L)]

ggplot(activity_profile, aes(x = bucket_5m, y = avg_events)) +
  geom_line(colour = "steelblue", linewidth = 0.7) +
  geom_smooth(se = FALSE, colour = "firebrick", linewidth = 0.5, linetype = "dashed") +
  scale_x_continuous(
    breaks = seq(0, 100, by = 12),
    labels = activity_profile$time_label[seq(1, nrow(activity_profile), by = 12)]
  ) +
  labs(
    title    = "Average intraday activity profile — April 2020",
    subtitle = "Average midquote-change events per 5-minute bucket across all trading days.\nPronounced diurnal pattern motivates seasonality adjustment in notebook 03.",
    x        = "London time",
    y        = "Avg events per 5-min bucket"
  ) +
  theme_minimal(base_size = 12)
```

```{r acf-durations}
# ACF of raw durations — tests whether consecutive durations are serially correlated.
# Significant autocorrelation (positive, decaying slowly) is the empirical
# justification for the ACD model: durations are NOT iid.
# If ACF were flat at zero, there would be no need for an ACD model.

# Use a sample for speed (full 1.26M series takes time); first 10,000 events
# captures the autocorrelation structure clearly
acf_result <- acf(
  evt_apr$duration[1:10000],
  lag.max = 50,
  plot    = FALSE
)

acf_df <- data.frame(
  lag = acf_result$lag[-1],
  acf = acf_result$acf[-1]
)

ci <- qnorm(0.975) / sqrt(10000)  # 95% confidence band

ggplot(acf_df, aes(x = lag, y = acf)) +
  geom_hline(yintercept = 0, colour = "black", linewidth = 0.4) +
  geom_hline(yintercept =  ci, colour = "firebrick", linetype = "dashed") +
  geom_hline(yintercept = -ci, colour = "firebrick", linetype = "dashed") +
  geom_segment(aes(xend = lag, yend = 0), colour = "steelblue") +
  geom_point(size = 1.5, colour = "steelblue") +
  labs(
    title    = "ACF of inter-event durations — April 2020 (first 10,000 events)",
    subtitle = "Significant positive autocorrelation confirms duration clustering.\nDashed lines = 95% confidence band under iid null.",
    x        = "Lag",
    y        = "Autocorrelation"
  ) +
  theme_minimal(base_size = 12)
```

```{r return-distribution}
# Return distribution vs normal — shows fat tails and excess kurtosis.
# This visually motivates the Student-t innovation distribution
# in the GARCH specifications in notebook 06.

ret_vals <- evt_apr$ret_mid
ret_mean <- mean(ret_vals)
ret_sd   <- sd(ret_vals)

ggplot(data.frame(ret = ret_vals), aes(x = ret)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 150, fill = "steelblue", alpha = 0.6, colour = "white") +
  stat_function(
    fun  = dnorm,
    args = list(mean = ret_mean, sd = ret_sd),
    colour = "firebrick", linewidth = 0.8
  ) +
  coord_cartesian(xlim = c(-0.003, 0.003)) +
  labs(
    title    = "Distribution of event-time log returns — April 2020",
    subtitle = sprintf("Red = fitted Normal(μ=%.2e, σ=%.2e). Heavy tails relative to normal motivate Student-t GARCH.",
                       ret_mean, ret_sd),
    x        = "Log return (event-time)",
    y        = "Density"
  ) +
  theme_minimal(base_size = 12)

# Print kurtosis — excess kurtosis >> 0 confirms fat tails
cat(sprintf("Excess kurtosis of April returns: %.2f (Normal = 0)\n",
            moments::kurtosis(ret_vals) - 3))
cat(sprintf("Skewness: %.4f\n", moments::skewness(ret_vals)))
```
```{r}
evt_apr[, bucket_5m := NULL]
```

---

## 12. Save Outputs

```{r save-outputs}
# We save the clean tick data and all three event-time series.
# Subsequent notebooks read these .rds files, so that they never re-read the raw CSV.
# .rds preserves all R object attributes (data.table keys, POSIXct timezone, etc.)

saveRDS(tick_clean, here::here("data", "processed", "tick_clean.rds"))
saveRDS(evt_apr,    here::here("data", "processed", "events_apr.rds"))
saveRDS(evt_may,    here::here("data", "processed", "events_may.rds"))
saveRDS(evt_jun,    here::here("data", "processed", "events_jun.rds"))

cat("Saved:\n")
cat("  data/processed/tick_clean.rds  —", format(nrow(tick_clean), big.mark=","), "rows\n")
cat("  data/processed/events_apr.rds  —", format(nrow(evt_apr),    big.mark=","), "rows\n")
cat("  data/processed/events_may.rds  —", format(nrow(evt_may),    big.mark=","), "rows\n")
cat("  data/processed/events_jun.rds  —", format(nrow(evt_jun),    big.mark=","), "rows\n")
```
