---
title: "03 — Cleaning & Intraday Seasonality Removal"
subtitle: "ACD-GARCH Intraday Volatility & Market Risk Forecasting — BP LSE Tick Data"
author: "Sourodeep Roy"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

## Purpose

This notebook prepares the event-time duration series for ACD modelling through
two steps:

1. **Upper-tail winsorisation** — caps extreme durations using April-calibrated
   bounds to stabilise parameter estimation without distorting the bulk of the
   distribution.

2. **Intraday seasonality removal** — removes the time-of-day (diurnal) pattern
   from durations using a non-parametric median profile estimated on April only
   and applied unchanged to May and June.

Both steps use **April-calibrated parameters only**. No information from May or
June is used in any cleaning or adjustment decision. This is essential for
maintaining a leakage-free backtest.

- **Inputs:**
  - `data/processed/events_apr.rds`
  - `data/processed/events_may.rds`
  - `data/processed/events_jun.rds`
- **Outputs:**
  - `data/processed/events_apr_clean.rds`
  - `data/processed/events_may_clean.rds`
  - `data/processed/events_jun_clean.rds`

---

## 1. Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5)
```

```{r load-packages}
suppressPackageStartupMessages({
  library(data.table)
  library(lubridate)
  library(ggplot2)
  library(here)
})
```

---

## 2. Load Data

```{r load-data}
evt_apr <- readRDS(here::here("data", "processed", "events_apr.rds"))
evt_may <- readRDS(here::here("data", "processed", "events_may.rds"))
evt_jun <- readRDS(here::here("data", "processed", "events_jun.rds"))

cat("Loaded:\n")
cat(sprintf("  events_apr: %s rows, %d trading days\n",
            format(nrow(evt_apr), big.mark=","), uniqueN(evt_apr$date)))
cat(sprintf("  events_may: %s rows, %d trading days\n",
            format(nrow(evt_may), big.mark=","), uniqueN(evt_may$date)))
cat(sprintf("  events_jun: %s rows, %d trading days\n",
            format(nrow(evt_jun), big.mark=","), uniqueN(evt_jun$date)))
```

```{r pre-checks}
# Verify the data is in the expected state before any cleaning.
# These checks ensure notebook 02 outputs haven't been corrupted.

required_cols <- c("timestamp", "date", "bid", "ask", "mid",
                   "ret_mid", "duration", "trading_seconds")

for (nm in c("evt_apr", "evt_may", "evt_jun")) {
  dt <- get(nm)
  missing <- setdiff(required_cols, names(dt))
  if (length(missing) > 0)
    stop(nm, " is missing columns: ", paste(missing, collapse = ", "))
}

# Durations must be strictly positive and finite going into cleaning
stopifnot(all(evt_apr$duration > 0), all(is.finite(evt_apr$duration)))
stopifnot(all(evt_apr$ask >= evt_apr$bid))

# Timestamps correctly ordered within days
n_reversals <- evt_apr[,
  sum(diff(as.numeric(timestamp)) < 0), by = date][, sum(V1)]
stopifnot("Timestamp reversals found in April" = n_reversals == 0)

cat("Pre-checks passed.\n")
cat("\nApril duration tail (before cleaning):\n")
print(quantile(evt_apr$duration,
               probs = c(0.95, 0.99, 0.995, 0.999, 0.9999)))
```

---

## 3. Upper-Tail Winsorisation

### Design rationale

Duration distributions in high-frequency data have a heavy right tail — long
gaps during lunch, after large news events, or at the end of the day. These
extreme values inflate the ACD log-likelihood and can destabilise parameter
estimation without representing genuine market microstructure dynamics.

We cap durations at the **99.5th percentile of April**. This removes the top
0.5% of the most extreme gaps while leaving the bulk of the distribution
unchanged. Crucially:

- The cap threshold is estimated on **April only**
- The same fixed threshold is applied to May and June
- We do **not** cap the lower tail — the ~50ms floor is a genuine
  market microstructure feature (LSE Minimum Quote Life), not an outlier

```{r winsorise}
# Estimate the upper cap on April training data only
upper_alpha <- 0.995
dur_cap <- as.numeric(quantile(evt_apr$duration, probs = upper_alpha,
                               na.rm = TRUE))

cat(sprintf("April upper duration cap (%.1f%% quantile): %.4f seconds\n",
            upper_alpha * 100, dur_cap))

# Apply the April-derived cap to all three months.
# pmin() replaces any duration exceeding the cap with the cap value.
# We create duration_w (winsorised) and log_duration_w as new columns,
# leaving the original duration column intact for reference.

apply_upper_cap <- function(dt, cap) {
  dt <- data.table::copy(dt)
  stopifnot(is.finite(cap), cap > 0)
  dt[, duration_w     := pmin(duration, cap)]
  dt[, log_duration_w := log(duration_w)]
  dt[]
}

evt_apr_clean <- apply_upper_cap(evt_apr, dur_cap)
evt_may_clean <- apply_upper_cap(evt_may, dur_cap)
evt_jun_clean <- apply_upper_cap(evt_jun, dur_cap)

# Report how many observations hit the cap in each month.
# For April this should be ~0.5% by construction.
# May/June may differ — that is expected and acceptable.
cap_report <- rbindlist(list(
  data.table(Month = "April", N = nrow(evt_apr_clean),
             Capped = sum(evt_apr_clean$duration_w == dur_cap)),
  data.table(Month = "May",   N = nrow(evt_may_clean),
             Capped = sum(evt_may_clean$duration_w == dur_cap)),
  data.table(Month = "June",  N = nrow(evt_jun_clean),
             Capped = sum(evt_jun_clean$duration_w == dur_cap))
))
cap_report[, Pct_Capped := round(100 * Capped / N, 3)]
print(cap_report)
```

```{r winsorise-plot}
# Compare raw vs winsorised duration distributions on log scale.
# The only difference should be visible at the far right tail.
# The bulk of the distribution should be identical.

plot_data <- rbind(
  data.table(duration = evt_apr$duration,         series = "Raw"),
  data.table(duration = evt_apr_clean$duration_w, series = "Winsorised")
)

ggplot(plot_data, aes(x = duration, fill = series)) +
  geom_histogram(bins = 120, alpha = 0.6, position = "identity",
                 colour = "white") +
  scale_x_log10() +
  scale_fill_manual(values = c("Raw" = "steelblue",
                               "Winsorised" = "firebrick")) +
  labs(
    title    = "April durations: raw vs winsorised (log scale)",
    subtitle = sprintf("Upper cap at %.4fs (99.5th percentile of April). Only far right tail affected.",
                       dur_cap),
    x        = "Duration (seconds, log scale)",
    y        = "Count",
    fill     = ""
  ) +
  theme_minimal(base_size = 12)
```

---

## 4. Intraday Seasonality Removal

### Why seasonality removal is necessary

The intraday activity profile in notebook 02 showed a clear U-shaped pattern:
high activity (short durations) at the open and close, and lower activity
(long durations) around midday. If this diurnal pattern is not removed, the
ACD model will partially fit the time-of-day pattern rather than the genuine
clustering dynamics we want to capture.

### Why non-parametric (NP) median profile over FFF

Two approaches are common in the literature:

**Flexible Fourier Form (FFF):** fits a parametric harmonic regression to the
diurnal shape. Requires assumptions about the functional form and number of
harmonics (tuning parameter Q).

**Non-parametric (NP) median profile:** computes the median duration within
each time-of-day bucket. No functional form assumptions. Robust to outliers
within each bucket (median vs mean). Simple and transparent.

We choose NP for three reasons:
1. No parametric assumptions about the shape of the diurnal curve
2. Median is robust — a few extreme durations in a bucket don't distort the profile
3. The April 2020 diurnal pattern (confirmed in notebook 02) is irregular enough
   that a parametric harmonic approximation would require many terms to capture
   the 14:30 US open spike

**Critical leakage note:** The NP profile is estimated on **April data only**.
The same fixed profile is then applied to May and June without any refitting.
Estimating a separate profile for May or June would use test-set information
to adjust the data — a form of data leakage that would overstate performance.

```{r add-tod-fields}
# Add time-of-day fields needed for the diurnal profile.
# We use 30-minute buckets (17 buckets across the 08:00-16:30 trading day).

add_tod_fields <- function(dt) {
  dt <- data.table::copy(dt)
  dt[, tod_min   := hour(timestamp) * 60L + minute(timestamp)]
  dt[, bucket_30m := floor(tod_min / 30L) * 30L]
  dt[, bucket_time := sprintf("%02d:%02d",
                              bucket_30m %/% 60L,
                              bucket_30m %% 60L)]
  dt[]
}

evt_apr_clean <- add_tod_fields(evt_apr_clean)
evt_may_clean <- add_tod_fields(evt_may_clean)
evt_jun_clean <- add_tod_fields(evt_jun_clean)
```

```{r np-profile}
# Estimate the NP diurnal profile on April only.
# Step 1: median winsorised duration within each 30-minute bucket
# Step 2: normalise scale factors to mean 1 so the adjusted series
#         has the same average duration as the original

apr_profile <- evt_apr_clean[, .(
  med_dur = median(duration_w, na.rm = TRUE),
  n_events = .N
), by = .(bucket_30m, bucket_time)][order(bucket_30m)]

# Normalise: scale_factor > 1 means the bucket is slow (long durations)
#            scale_factor < 1 means the bucket is fast (short durations)
apr_profile[, scale_factor := med_dur / mean(med_dur, na.rm = TRUE)]

cat("April NP diurnal profile (30-min buckets):\n")
print(apr_profile[, .(bucket_time, med_dur = round(med_dur, 4),
                      scale_factor = round(scale_factor, 4), n_events)])
```

```{r apply-np-profile}
# Apply the April-calibrated profile to all three months.
# duration_adj_np = duration_w / scale_factor
# This divides out the expected diurnal component, leaving only
# the genuine clustering dynamics for the ACD model to capture.
#
# We use an in-place join to add scale_factor without reordering columns.

apply_np_adjustment <- function(dt, profile) {
  dt <- data.table::copy(dt)

  # In-place join: add scale_factor from profile matched on bucket_30m
  dt[profile, on = "bucket_30m", scale_factor := i.scale_factor]

  # If any bucket has no match (shouldn't happen, but defensive)
  n_missing <- sum(is.na(dt$scale_factor))
  if (n_missing > 0) {
    warning(n_missing, " rows have no diurnal profile match — scale_factor set to 1")
    dt[is.na(scale_factor), scale_factor := 1]
  }

  dt[, duration_adj_np    := duration_w / scale_factor]
  dt[, log_duration_adj_np := log(duration_adj_np)]
  dt[]
}

evt_apr_clean <- apply_np_adjustment(evt_apr_clean, apr_profile)
evt_may_clean <- apply_np_adjustment(evt_may_clean, apr_profile)
evt_jun_clean <- apply_np_adjustment(evt_jun_clean, apr_profile)

cat("NP adjustment applied to April, May, June using April-calibrated profile.\n")
```

```{r diurnal-before-after}
# Visual check: diurnal pattern before and after NP adjustment on April.
# If adjustment worked, the time-of-day variation should largely flatten.

diurnal_compare <- evt_apr_clean[, .(
  Before = mean(duration_w,       na.rm = TRUE),
  After  = mean(duration_adj_np,  na.rm = TRUE)
), by = .(bucket_30m, bucket_time)][order(bucket_30m)]

diurnal_long <- melt(
  diurnal_compare,
  id.vars      = c("bucket_30m", "bucket_time"),
  measure.vars = c("Before", "After"),
  variable.name = "Series",
  value.name    = "Mean_Duration"
)

ggplot(diurnal_long,
       aes(x = bucket_time, y = Mean_Duration,
           group = Series, linetype = Series, colour = Series)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c("Before" = "steelblue",
                                 "After"  = "firebrick")) +
  labs(
    title    = "April diurnal pattern: before vs after NP adjustment",
    subtitle = "After adjustment the time-of-day variation should be substantially reduced.",
    x        = "London time (30-min buckets)",
    y        = "Mean duration (seconds)",
    linetype = "", colour = ""
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r variance-reduction}
# Numeric summary of the variance reduction achieved by seasonality removal.
# A good diurnal adjustment should meaningfully reduce log-duration variance.
# We report variance of log-durations (standard in the ACD literature).

var_raw <- var(evt_apr_clean$log_duration_w,       na.rm = TRUE)
var_adj <- var(evt_apr_clean$log_duration_adj_np,  na.rm = TRUE)

var_table <- data.table(
  Series              = c("Log(winsorised duration)",
                          "Log(NP-adjusted duration)"),
  Variance            = round(c(var_raw, var_adj), 6),
  Reduction_vs_Raw_pct = round(100 * (1 - c(var_raw, var_adj) / var_raw), 2)
)
print(var_table)
```

---

## 5. ACF After Adjustment

```{r acf-comparison}
# We compare the ACF of log-durations before and after NP adjustment.
# Goal: confirm that the diurnal pattern has been removed from the
# autocorrelation structure, and that genuine duration clustering remains.
#
# If the pre-adjustment ACF was driven purely by the diurnal pattern,
# it would disappear after adjustment. If it persists (as we expect),
# it confirms genuine ACD-type clustering that the model needs to capture.

compute_acf_df <- function(x, lag_max = 50, n_label) {
  a <- acf(x[1:min(10000, length(x))], lag.max = lag_max, plot = FALSE)
  data.table(
    lag    = as.integer(a$lag[-1]),
    acf    = as.numeric(a$acf[-1]),
    series = n_label
  )
}

acf_before <- compute_acf_df(evt_apr_clean$log_duration_w,      lag_max=50, "Before adjustment")
acf_after  <- compute_acf_df(evt_apr_clean$log_duration_adj_np, lag_max=50, "After adjustment")
acf_all    <- rbind(acf_before, acf_after)

ci <- qnorm(0.975) / sqrt(10000)

ggplot(acf_all, aes(x = lag, y = acf, colour = series)) +
  geom_hline(yintercept = 0,   colour = "black",    linewidth = 0.3) +
  geom_hline(yintercept =  ci, colour = "firebrick", linetype = "dashed") +
  geom_hline(yintercept = -ci, colour = "firebrick", linetype = "dashed") +
  geom_segment(aes(xend = lag, yend = 0),
               position = position_dodge(width = 0.5)) +
  geom_point(size = 1.5,
             position = position_dodge(width = 0.5)) +
  facet_wrap(~series, ncol = 1) +
  scale_colour_manual(values = c("Before adjustment" = "steelblue",
                                 "After adjustment"  = "firebrick")) +
  labs(
    title    = "ACF of log-durations before and after NP adjustment — April 2020",
    subtitle = "Clustering autocorrelation should persist after adjustment (justifies ACD model).\nDashed lines = 95% CI under iid null.",
    x = "Lag", y = "Autocorrelation",
    colour = ""
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```

---

## 6. Final Series Selection

```{r final-series}
# We designate NP-adjusted log-durations as the modelling series.
# This is the series fed into the Weibull ACD(2,2) in notebook 04.

# Assign clean, unambiguous column names for downstream use
evt_apr_clean[, duration_final     := duration_adj_np]
evt_may_clean[, duration_final     := duration_adj_np]
evt_jun_clean[, duration_final     := duration_adj_np]

evt_apr_clean[, log_duration_final := log_duration_adj_np]
evt_may_clean[, log_duration_final := log_duration_adj_np]
evt_jun_clean[, log_duration_final := log_duration_adj_np]

# Final summary of the modelling series
cat("Final modelling series (duration_final) — April summary:\n")
print(summary(evt_apr_clean$duration_final))
cat(sprintf("\nLog-duration variance after full cleaning: %.6f\n",
            var(evt_apr_clean$log_duration_final)))
```

---

## 7. Save Outputs

```{r save-outputs}
saveRDS(evt_apr_clean, here::here("data", "processed", "events_apr_clean.rds"))
saveRDS(evt_may_clean, here::here("data", "processed", "events_may_clean.rds"))
saveRDS(evt_jun_clean, here::here("data", "processed", "events_jun_clean.rds"))

# Also save the April diurnal profile — needed if we want to inspect or
# reproduce the adjustment without re-running this notebook
saveRDS(apr_profile, here::here("data", "processed", "apr_diurnal_profile.rds"))

cat("Saved:\n")
cat("  data/processed/events_apr_clean.rds —",
    format(nrow(evt_apr_clean), big.mark=","), "rows\n")
cat("  data/processed/events_may_clean.rds —",
    format(nrow(evt_may_clean), big.mark=","), "rows\n")
cat("  data/processed/events_jun_clean.rds —",
    format(nrow(evt_jun_clean), big.mark=","), "rows\n")
cat("  data/processed/apr_diurnal_profile.rds — April NP profile\n")
```
