---
title: "03 — Cleaning & Intraday Seasonality Removal"
subtitle: "ACD-GARCH Intraday Volatility & Market Risk Forecasting — BP LSE Tick Data"
author: "Sourodeep Roy"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

## Purpose

This notebook prepares the event-time duration series for ACD modelling through
two steps:

1. **Upper-tail winsorisation** — caps extreme durations using April-calibrated
   bounds to stabilise parameter estimation without distorting the bulk of the
   distribution.

2. **Intraday seasonality removal** — removes the time-of-day (diurnal) pattern
   from durations using a non-parametric median profile estimated on April only
   and applied unchanged to May and June.

Both steps use **April-calibrated parameters only**. No information from May or
June is used in any cleaning or adjustment decision. This is essential for
maintaining a leakage-free backtest.

- **Inputs:**
  - `data/processed/events_apr.rds`
  - `data/processed/events_may.rds`
  - `data/processed/events_jun.rds`
- **Outputs:**
  - `data/processed/events_apr_clean.rds`
  - `data/processed/events_may_clean.rds`
  - `data/processed/events_jun_clean.rds`

---

## 1. Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5)
```

```{r load-packages}
suppressPackageStartupMessages({
  library(data.table)
  library(lubridate)
  library(ggplot2)
  library(here)
})
```

---

## 2. Load Data

```{r load-data}
evt_apr <- readRDS(here::here("data", "processed", "events_apr.rds"))
evt_may <- readRDS(here::here("data", "processed", "events_may.rds"))
evt_jun <- readRDS(here::here("data", "processed", "events_jun.rds"))

cat("Loaded:\n")
cat(sprintf("  events_apr: %s rows, %d trading days\n",
            format(nrow(evt_apr), big.mark=","), uniqueN(evt_apr$date)))
cat(sprintf("  events_may: %s rows, %d trading days\n",
            format(nrow(evt_may), big.mark=","), uniqueN(evt_may$date)))
cat(sprintf("  events_jun: %s rows, %d trading days\n",
            format(nrow(evt_jun), big.mark=","), uniqueN(evt_jun$date)))
```

```{r pre-checks}
# Verify the data is in the expected state before any cleaning.
# These checks ensure notebook 02 outputs haven't been corrupted.

required_cols <- c("timestamp", "date", "bid", "ask", "mid",
                   "ret_mid", "duration", "trading_seconds")

for (nm in c("evt_apr", "evt_may", "evt_jun")) {
  dt <- get(nm)
  missing <- setdiff(required_cols, names(dt))
  if (length(missing) > 0)
    stop(nm, " is missing columns: ", paste(missing, collapse = ", "))
}

# Durations must be strictly positive and finite going into cleaning
stopifnot(all(evt_apr$duration > 0), all(is.finite(evt_apr$duration)))
stopifnot(all(evt_apr$ask >= evt_apr$bid))

# Timestamps correctly ordered within days
n_reversals <- evt_apr[,
  sum(diff(as.numeric(timestamp)) < 0), by = date][, sum(V1)]
stopifnot("Timestamp reversals found in April" = n_reversals == 0)

cat("Pre-checks passed.\n")
cat("\nApril duration tail (before cleaning):\n")
print(quantile(evt_apr$duration,
               probs = c(0.95, 0.99, 0.995, 0.999, 0.9999)))
```

---

## 3. Upper-Tail Winsorisation

### Design rationale

Duration distributions in high-frequency data have a heavy right tail — long
gaps during lunch, after large news events, or at the end of the day. These
extreme values inflate the ACD log-likelihood and can destabilise parameter
estimation without representing genuine market microstructure dynamics.

We cap durations at the **99.5th percentile of April**. This removes the top
0.5% of the most extreme gaps while leaving the bulk of the distribution
unchanged. Crucially:

- The cap threshold is estimated on **April only**
- The same fixed threshold is applied to May and June
- We do **not** cap the lower tail — the ~50ms floor is a genuine
  market microstructure feature (LSE Minimum Quote Life), not an outlier

```{r winsorise}
# Estimate the upper cap on April training data only
upper_alpha <- 0.995
dur_cap <- as.numeric(quantile(evt_apr$duration, probs = upper_alpha,
                               na.rm = TRUE))

cat(sprintf("April upper duration cap (%.1f%% quantile): %.4f seconds\n",
            upper_alpha * 100, dur_cap))

# Apply the April-derived cap to all three months.
# pmin() replaces any duration exceeding the cap with the cap value.
# We create duration_w (winsorised) and log_duration_w as new columns,
# leaving the original duration column intact for reference.

apply_upper_cap <- function(dt, cap) {
  dt <- data.table::copy(dt)
  stopifnot(is.finite(cap), cap > 0)
  dt[, duration_w     := pmin(duration, cap)]
  dt[, log_duration_w := log(duration_w)]
  dt[]
}

evt_apr_clean <- apply_upper_cap(evt_apr, dur_cap)
evt_may_clean <- apply_upper_cap(evt_may, dur_cap)
evt_jun_clean <- apply_upper_cap(evt_jun, dur_cap)

# Report how many observations hit the cap in each month.
# For April this should be ~0.5% by construction.
# May/June may differ — that is expected and acceptable.
cap_report <- rbindlist(list(
  data.table(Month = "April", N = nrow(evt_apr_clean),
             Capped = sum(evt_apr_clean$duration_w == dur_cap)),
  data.table(Month = "May",   N = nrow(evt_may_clean),
             Capped = sum(evt_may_clean$duration_w == dur_cap)),
  data.table(Month = "June",  N = nrow(evt_jun_clean),
             Capped = sum(evt_jun_clean$duration_w == dur_cap))
))
cap_report[, Pct_Capped := round(100 * Capped / N, 3)]
print(cap_report)
```
# May's higher capping rate (1.128%) reflects lower overall market activity
# in May 2020 as volatility subsided after the April COVID crash peak.
# Fewer events + longer quiet periods = more durations exceeding the
# April-calibrated cap. This is expected and does not indicate a problem.
# The fixed April cap is deliberately conservative for test-month application.

```{r winsorise-plot}
# Compare raw vs winsorised duration distributions on log scale.
# The only difference should be visible at the far right tail.
# The bulk of the distribution should be identical.

plot_data <- rbind(
  data.table(duration = evt_apr$duration,         series = "Raw"),
  data.table(duration = evt_apr_clean$duration_w, series = "Winsorised")
)

ggplot(plot_data, aes(x = duration, fill = series)) +
  geom_histogram(bins = 120, alpha = 0.6, position = "identity",
                 colour = "white") +
  scale_x_log10() +
  scale_fill_manual(values = c("Raw" = "steelblue",
                               "Winsorised" = "firebrick")) +
  labs(
    title    = "April durations: raw vs winsorised (log scale)",
    subtitle = sprintf("Upper cap at %.4fs (99.5th percentile of April). Only far right tail affected.",
                       dur_cap),
    x        = "Duration (seconds, log scale)",
    y        = "Count",
    fill     = ""
  ) +
  theme_minimal(base_size = 12)
```

---

## 4. Intraday Seasonality Removal

### Why seasonality removal is necessary

The intraday activity profile in notebook 02 showed a clear U-shaped pattern:
high activity (short durations) at the open and close, and lower activity
(long durations) around midday. If this diurnal pattern is not removed, the
ACD model will partially fit the time-of-day pattern rather than the genuine
clustering dynamics we want to capture.

### Why non-parametric (NP) median profile over FFF

Two approaches are common in the literature:

**Flexible Fourier Form (FFF):** fits a parametric harmonic regression to the
diurnal shape. Requires assumptions about the functional form and number of
harmonics (tuning parameter Q).

**Non-parametric (NP) median profile:** computes the median duration within
each time-of-day bucket. No functional form assumptions. Robust to outliers
within each bucket (median vs mean). Simple and transparent.

We choose NP for three reasons:
1. No parametric assumptions about the shape of the diurnal curve
2. Median is robust — a few extreme durations in a bucket don't distort the profile
3. The April 2020 diurnal pattern (confirmed in notebook 02) is irregular enough
   that a parametric harmonic approximation would require many terms to capture
   the 14:30 US open spike

**Critical leakage note:** The NP profile is estimated on **April data only**.
The same fixed profile is then applied to May and June without any refitting.
Estimating a separate profile for May or June would use test-set information
to adjust the data — a form of data leakage that would overstate performance.

```{r add-tod-fields}
# Time-of-day bucket assignment using 5-minute intervals.
# bucket_5m is computed RELATIVE TO MARKET OPEN (08:00 = bucket 0).
# This ensures:
#   bucket 0   = 08:00–08:05
#   bucket 17  = 09:25–09:30
#   bucket 101 = 16:25–16:30
# The relative indexing keeps bucket numbers intuitive and ensures the
# label formula produces correct clock times.

add_tod_fields <- function(dt) {
  dt <- data.table::copy(dt)

  # Total seconds since midnight
  dt[, tod_sec := hour(timestamp) * 3600L +
                  minute(timestamp) * 60L +
                  as.integer(second(timestamp))]

  # Seconds since market open (08:00:00 = 28800 seconds since midnight)
  open_sec <- 8L * 3600L   # 28800

  # 5-minute bucket relative to open: 0 = 08:00, 1 = 08:05, ..., 101 = 16:25
  dt[, bucket_5m := (tod_sec - open_sec) %/% 300L]

  # Readable clock label: bucket 0 = "08:00", bucket 12 = "09:00", etc.
  dt[, bucket_time := sprintf("%02d:%02d",
                               8L + (bucket_5m * 5L) %/% 60L,
                               (bucket_5m * 5L) %% 60L)]

  # Drop intermediate column
  dt[, tod_sec := NULL]
  dt[]
}

evt_apr_clean <- add_tod_fields(evt_apr_clean)
evt_may_clean <- add_tod_fields(evt_may_clean)
evt_jun_clean <- add_tod_fields(evt_jun_clean)

# Verify bucket range is 0–101 (102 buckets across 8.5-hour trading day)
cat("April bucket_5m range:", min(evt_apr_clean$bucket_5m),
    "to", max(evt_apr_clean$bucket_5m), "\n")
cat("Expected: 0 to 101\n")

# Spot check a few known times
cat("\nSpot checks:\n")
cat("First event bucket_time:", evt_apr_clean$bucket_time[1], "(should be 08:xx)\n")
```


```{r np-profile}
# NP median profile using 5-minute buckets.
# Finer granularity is necessary here because the 14:30 US market open
# spike is sharp and confined to a narrow window — a 30-minute bucket
# averages over both the spike and its decay, leaving substantial
# within-bucket diurnal variation unexplained.
# 5-minute buckets give 102 intervals across the 08:00-16:30 trading day.

apr_profile <- evt_apr_clean[, .(
  med_dur  = median(duration_w, na.rm = TRUE),
  n_events = .N
), by = .(bucket_5m, bucket_time)][order(bucket_5m)]

apr_profile[, scale_factor := med_dur / mean(med_dur, na.rm = TRUE)]

cat("Profile buckets:", nrow(apr_profile), "\n")
cat("Min scale factor:", round(min(apr_profile$scale_factor), 4),
    "at", apr_profile[which.min(scale_factor), bucket_time], "\n")
cat("Max scale factor:", round(max(apr_profile$scale_factor), 4),
    "at", apr_profile[which.max(scale_factor), bucket_time], "\n")

print(apr_profile[, .(bucket_time,
                      med_dur      = round(med_dur, 4),
                      scale_factor = round(scale_factor, 4),
                      n_events)])
```


```{r apply-np-adjustment}
apply_np_adjustment <- function(dt, profile) {
  dt <- data.table::copy(dt)

  dt[profile, on = "bucket_5m", scale_factor := i.scale_factor]

  n_missing <- sum(is.na(dt$scale_factor))
  if (n_missing > 0) {
    warning(n_missing, " rows have no diurnal profile match — scale_factor set to 1")
    dt[is.na(scale_factor), scale_factor := 1]
  }

  dt[, duration_adj_np     := duration_w / scale_factor]
  dt[, log_duration_adj_np := log(duration_adj_np)]
  dt[]
}

evt_apr_clean <- apply_np_adjustment(evt_apr_clean, apr_profile)
evt_may_clean <- apply_np_adjustment(evt_may_clean, apr_profile)
evt_jun_clean <- apply_np_adjustment(evt_jun_clean, apr_profile)

cat("NP adjustment applied to April, May, June using April-calibrated profile.\n")
```


```{r diurnal-before-after}
# Visual check: diurnal pattern before and after NP adjustment on April.
# If adjustment worked, the time-of-day variation should largely flatten.
# X-axis shows hourly labels only (every 12th bucket) to keep plot readable.

diurnal_compare <- evt_apr_clean[, .(
  Before = mean(duration_w,      na.rm = TRUE),
  After  = mean(duration_adj_np, na.rm = TRUE)
), by = .(bucket_5m, bucket_time)][order(bucket_5m)]

diurnal_long <- melt(
  diurnal_compare,
  id.vars       = c("bucket_5m", "bucket_time"),
  measure.vars  = c("Before", "After"),
  variable.name = "Series",
  value.name    = "Mean_Duration"
)

# Show only hourly tick marks (every 12 buckets = 60 minutes)
# to avoid x-axis overcrowding with 102 labels
hourly_breaks  <- diurnal_compare$bucket_5m[seq(1, nrow(diurnal_compare), by = 12)]
hourly_labels  <- diurnal_compare$bucket_time[seq(1, nrow(diurnal_compare), by = 12)]

ggplot(diurnal_long,
       aes(x = bucket_5m, y = Mean_Duration,
           group = Series, linetype = Series, colour = Series)) +
  geom_line(linewidth = 0.8) +
  scale_x_continuous(breaks = hourly_breaks, labels = hourly_labels) +
  scale_colour_manual(values = c("Before" = "steelblue",
                                 "After"  = "firebrick")) +
  labs(
    title    = "April diurnal pattern: before vs after NP adjustment",
    subtitle = "After adjustment the time-of-day variation should be substantially reduced.",
    x        = "London time (5-min buckets, hourly labels)",
    y        = "Mean duration (seconds)",
    linetype = "", colour = ""
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r variance-reduction}
# Numeric summary of the variance reduction achieved by seasonality removal.
# A good diurnal adjustment should meaningfully reduce log-duration variance.
# We report variance of log-durations (standard in the ACD literature).
# With 5-minute buckets we expect 10-30% reduction vs the 1.6% seen with
# 30-minute buckets.
# NOTE: 1.76% variance reduction is low relative to typical ACD literature values
# (10-30%). This reflects the data structure rather than a methodological failure:
# within-bucket variance is dominated by (1) the ~50ms LSE MQL floor creating
# a mass point at log(0.051) ≈ -3.0, and (2) a heavy right tail extending to
# log(5.84) ≈ 1.76 — a 4.76 log-unit within-bucket range. The diurnal component
# (median ratio ~2x between slowest/fastest buckets) is small relative to this
# within-bucket variation. The NP adjustment correctly removes the between-bucket
# diurnal structure; the ACF comparison below confirms genuine clustering remains.
var_raw <- var(evt_apr_clean$log_duration_w,      na.rm = TRUE)
var_adj <- var(evt_apr_clean$log_duration_adj_np, na.rm = TRUE)

var_table <- data.table(
  Series               = c("Log(winsorised duration)",
                            "Log(NP-adjusted, 5-min buckets)"),
  Variance             = round(c(var_raw, var_adj), 6),
  Reduction_vs_Raw_pct = round(100 * (1 - c(var_raw, var_adj) / var_raw), 2)
)
print(var_table)
```

---

## 5. ACF After Adjustment

```{r acf-comparison}
# We compare ACF of log-durations before and after NP adjustment.
# Two questions we are answering:
#
# Q1: Does the diurnal pattern drive the autocorrelation?
#     If yes: ACF would collapse after adjustment → ACD model less justified
#     If no:  ACF persists after adjustment → genuine clustering → ACD needed
#
# Q2: Is the adjustment doing something useful despite low variance reduction?
#     If ACF structure is similar before/after, the adjustment has correctly
#     removed the diurnal component while preserving the clustering signal.
#
# We use the first 10,000 events for speed. This is sufficient to estimate
# ACF structure at lags 1-50 with tight confidence intervals.

compute_acf_df <- function(x, lag_max = 50, n_label) {
  a <- acf(x[1:min(10000, length(x))], lag.max = lag_max, plot = FALSE)
  data.table(
    lag    = as.integer(a$lag[-1]),
    acf    = as.numeric(a$acf[-1]),
    series = n_label
  )
}

acf_before <- compute_acf_df(evt_apr_clean$log_duration_w,
                              lag_max = 50, "Before NP adjustment")
acf_after  <- compute_acf_df(evt_apr_clean$log_duration_adj_np,
                              lag_max = 50, "After NP adjustment")
acf_all    <- rbind(acf_before, acf_after)

ci <- qnorm(0.975) / sqrt(10000)

ggplot(acf_all, aes(x = lag, y = acf, colour = series)) +
  geom_hline(yintercept = 0,    colour = "black",    linewidth = 0.3) +
  geom_hline(yintercept =  ci,  colour = "firebrick", linetype = "dashed") +
  geom_hline(yintercept = -ci,  colour = "firebrick", linetype = "dashed") +
  geom_segment(aes(xend = lag, yend = 0),
               position = position_dodge(width = 0.5)) +
  geom_point(size = 1.5,
             position = position_dodge(width = 0.5)) +
  facet_wrap(~series, ncol = 1) +
  scale_colour_manual(values = c("Before NP adjustment" = "steelblue",
                                 "After NP adjustment"  = "firebrick")) +
  labs(
    title    = "ACF of log-durations before and after NP adjustment — April 2020",
    subtitle = paste(
      "Persistent autocorrelation after adjustment confirms genuine duration clustering.",
      "Dashed lines = 95% CI under iid null. Sample = first 10,000 events.",
      sep = "\n"
    ),
    x = "Lag", y = "Autocorrelation", colour = ""
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")

# Print numeric summary for the top 5 lags — useful for the dissertation table
cat("\nACF at lags 1-5 (before adjustment):\n")
print(acf_before[lag <= 5, .(lag, acf = round(acf, 4))])
cat("\nACF at lags 1-5 (after adjustment):\n")
print(acf_after[lag <= 5,  .(lag, acf = round(acf, 4))])
```
# NOTE: Negative lag-1 ACF (-0.086 before, -0.099 after) is a known
# microstructure artefact of the ~50ms LSE MQL floor. When a duration
# hits the floor, the next duration tends to revert to a longer value,
# producing mechanical negative lag-1 autocorrelation. The ACD(2,2)
# MA component captures this pattern — it is not a modelling concern.

```{r ljung-box}
# Ljung-Box test for remaining autocorrelation in log-durations after adjustment.
# H0: the series is iid (no autocorrelation up to lag K)
# We test at lag 10 and lag 20.
# We EXPECT to reject H0 — this confirms the ACD model is necessary.
# If we failed to reject, the adjusted series would be iid and ACD would add nothing.

lb_before <- Box.test(evt_apr_clean$log_duration_w[1:10000],
                      lag = 20, type = "Ljung-Box")
lb_after  <- Box.test(evt_apr_clean$log_duration_adj_np[1:10000],
                      lag = 20, type = "Ljung-Box")

lb_table <- data.table(
  Series    = c("Before NP adjustment", "After NP adjustment"),
  Statistic = round(c(lb_before$statistic, lb_after$statistic), 2),
  P_value   = format(c(lb_before$p.value, lb_after$p.value), scientific = TRUE),
  Decision  = c(
    ifelse(lb_before$p.value < 0.05, "Reject H0 — autocorrelation present", "Fail to reject"),
    ifelse(lb_after$p.value  < 0.05, "Reject H0 — autocorrelation present", "Fail to reject")
  )
)

cat("Ljung-Box test (lag = 20):\n")
print(lb_table)
cat("\nInterpretation: rejection of H0 after adjustment confirms genuine duration\n")
cat("clustering that is NOT an artefact of the diurnal pattern.\n")
cat("This is the formal statistical justification for the ACD model.\n")
```
```{r cleaning-summary}
# Compact summary of all cleaning decisions for the dissertation and README.
# Every parameter here was estimated on April only and applied to all months.

cleaning_summary <- data.table(
  Step = c(
    "Upper-tail winsorisation",
    "Diurnal adjustment method",
    "Bucket granularity",
    "Lower-tail treatment",
    "Profile estimation period",
    "Profile application"
  ),
  Decision = c(
    sprintf("Cap at 99.5th percentile of April = %.4f seconds", dur_cap),
    "Non-parametric median profile (NP)",
    "5-minute buckets (102 intervals across 08:00-16:30)",
    "No lower-tail capping — 50ms MQL floor is genuine market microstructure",
    "April 2020 only (training set)",
    "Fixed April profile applied unchanged to May and June"
  ),
  Rationale = c(
    "Stabilises ACD likelihood without distorting bulk of distribution",
    "No parametric assumptions; robust to within-bucket outliers",
    "Captures sharp 14:30 US open spike; 30-min buckets too coarse",
    "50ms floor reflects LSE Minimum Quote Life — not an outlier",
    "Prevents test-set information from contaminating cleaning step",
    "Ensures leakage-free backtest — no refitting on test months"
  )
)

print(cleaning_summary)
```

---

## 6. Final Series Selection

```{r final-series}
# We designate NP-adjusted log-durations as the modelling series.
# This is the series fed into the Weibull ACD(2,2) in notebook 04.

# Assign clean, unambiguous column names for downstream use
evt_apr_clean[, duration_final     := duration_adj_np]
evt_may_clean[, duration_final     := duration_adj_np]
evt_jun_clean[, duration_final     := duration_adj_np]

evt_apr_clean[, log_duration_final := log_duration_adj_np]
evt_may_clean[, log_duration_final := log_duration_adj_np]
evt_jun_clean[, log_duration_final := log_duration_adj_np]

# Final summary of the modelling series
cat("Final modelling series (duration_final) — April summary:\n")
print(summary(evt_apr_clean$duration_final))
cat(sprintf("\nLog-duration variance after full cleaning: %.6f\n",
            var(evt_apr_clean$log_duration_final)))
```

---

## 7. Save Outputs

```{r save-outputs}
saveRDS(evt_apr_clean, here::here("data", "processed", "events_apr_clean.rds"))
saveRDS(evt_may_clean, here::here("data", "processed", "events_may_clean.rds"))
saveRDS(evt_jun_clean, here::here("data", "processed", "events_jun_clean.rds"))

# Also save the April diurnal profile — needed if we want to inspect or
# reproduce the adjustment without re-running this notebook
saveRDS(apr_profile, here::here("data", "processed", "apr_diurnal_profile.rds"))

cat("Saved:\n")
cat("  data/processed/events_apr_clean.rds —",
    format(nrow(evt_apr_clean), big.mark=","), "rows\n")
cat("  data/processed/events_may_clean.rds —",
    format(nrow(evt_may_clean), big.mark=","), "rows\n")
cat("  data/processed/events_jun_clean.rds —",
    format(nrow(evt_jun_clean), big.mark=","), "rows\n")
cat("  data/processed/apr_diurnal_profile.rds — April NP profile\n")
```
