---
title: "05 — Intensity-Adjusted Returns and 5-Minute Aggregation"
subtitle: "ACD-GARCH Intraday Volatility & Market Risk Forecasting — BP LSE Tick Data"
author: "Sourodeep Roy"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: flatly
    code_folding: show
---

## Purpose

This notebook constructs intensity-adjusted (IA) returns from the event-level
tick data and aggregates them to a 5-minute frequency for GARCH estimation.

**The intensity adjustment** scales each event-level return by √(ψ̂_t), where
ψ̂_t is the conditional mean duration from the Weibull ACD(2,2) model fitted in
notebook 04. The intuition: a return observed during a fast-trading period
(small ψ̂_t, high intensity) carries less information per unit time than the same
return during a slow period. Scaling by √(ψ̂_t) equalises the information content
across events with different arrival rates, removing the activity-driven
heteroskedasticity before GARCH modelling.

**Two 5-minute return tracks are produced:**

- `r_ia_5m_sum` — sum of event-level IA returns within each 5-minute bucket.
  Used as the GARCH input (volatility track). Not directly invertible to raw
  returns, but correctly inherits the intensity-adjustment properties.

- `r_ia_5m_inv` — raw 5-minute return × √(mean ψ̂ in bucket). Invertible:
  dividing by √(ψ̂_5m) recovers the raw 5-minute return. Used for VaR/ES
  backtesting (risk track) where we need to map back to actual P&L.

**Key design principle:** All parameters (ψ̂_t values) come from notebook 04
and were estimated on April only. No new estimation occurs here.

- **Inputs:**
  - `data/processed/events_apr_model.rds`
  - `data/processed/events_may_model.rds`
  - `data/processed/events_jun_model.rds`
- **Outputs:**
  - `data/processed/r5_apr.rds` — April 5-minute return series
  - `data/processed/r5_may.rds` — May 5-minute return series
  - `data/processed/r5_jun.rds` — June 5-minute return series

---

## 1. Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5)
```

```{r load-packages}
suppressPackageStartupMessages({
  library(data.table)
  library(lubridate)
  library(ggplot2)
  library(here)
})
```

---

## 2. Load Data

```{r load-data}
# Load event-level data with psi_hat and lambda_hat already attached.
# These were produced and saved by notebook 04 — we do not re-run the ACD
# model or re-extract parameters here.

evt_apr_model <- readRDS(here::here("data", "processed", "events_apr_model.rds"))
evt_may_model <- readRDS(here::here("data", "processed", "events_may_model.rds"))
evt_jun_model <- readRDS(here::here("data", "processed", "events_jun_model.rds"))

# Verify psi_hat and lambda_hat are present
stopifnot("psi_hat"    %in% names(evt_apr_model),
          "lambda_hat" %in% names(evt_apr_model))
stopifnot("psi_hat"    %in% names(evt_may_model),
          "lambda_hat" %in% names(evt_may_model))
stopifnot("psi_hat"    %in% names(evt_jun_model),
          "lambda_hat" %in% names(evt_jun_model))

cat("Loaded:\n")
cat(sprintf("  April: %s events | psi_hat range: [%.4f, %.4f]\n",
            format(nrow(evt_apr_model), big.mark = ","),
            min(evt_apr_model$psi_hat), max(evt_apr_model$psi_hat)))
cat(sprintf("  May:   %s events | psi_hat range: [%.4f, %.4f]\n",
            format(nrow(evt_may_model), big.mark = ","),
            min(evt_may_model$psi_hat), max(evt_may_model$psi_hat)))
cat(sprintf("  June:  %s events | psi_hat range: [%.4f, %.4f]\n",
            format(nrow(evt_jun_model), big.mark = ","),
            min(evt_jun_model$psi_hat), max(evt_jun_model$psi_hat)))
```

---

## 3. Intensity-Adjusted Returns

```{r intensity-adjusted-returns}
# Intensity-adjusted return:
#   ret_ia_t = ret_mid_t * sqrt(psi_hat_t)
#
# Derivation: in event time, returns are measured over unequal calendar-time
# intervals. If we assume variance scales linearly with duration (as in the
# subordinated process framework of Clark 1973 and Engle & Russell 1998),
# then:
#   Var(ret_mid_t) ∝ psi_hat_t
# Dividing by sqrt(psi_hat_t) gives a series with approximately constant
# variance across events — removing the activity-driven heteroskedasticity
# that would otherwise contaminate GARCH estimation.
#
# NOTE: we scale by sqrt(psi_hat), not 1/sqrt(psi_hat).
# Higher psi_hat = longer expected duration = more time elapsed = more
# information content per return = larger scaled return. This is intentional.

evt_apr_model[, ret_ia := ret_mid * sqrt(psi_hat)]
evt_may_model[, ret_ia := ret_mid * sqrt(psi_hat)]
evt_jun_model[, ret_ia := ret_mid * sqrt(psi_hat)]

stopifnot(all(is.finite(evt_apr_model$ret_ia)))
stopifnot(all(is.finite(evt_may_model$ret_ia)))
stopifnot(all(is.finite(evt_jun_model$ret_ia)))

cat("Return summaries — raw vs intensity-adjusted:\n\n")
for (nm in c("Apr", "May", "Jun")) {
  dt <- switch(nm, Apr = evt_apr_model, May = evt_may_model, Jun = evt_jun_model)
  cat(sprintf("%s raw ret_mid:\n", nm)); print(summary(dt$ret_mid))
  cat(sprintf("%s adjusted ret_ia:\n", nm)); print(summary(dt$ret_ia))
  cat("\n")
}
```

---

## 4. Mean-Centering

```{r mean-centering}
# GARCH models assume a zero-mean innovation process.
# We subtract the sample mean of each series to enforce this.
# For intraday returns the mean is extremely close to zero, so this
# step has negligible numerical impact — but it is good practice.
#
# We centre each month by its own mean.
# For April this is in-sample centering.
# For May and June: since the intraday mean is ~0 by construction
# (no persistent drift over 6.5 trading hours), using the test-month
# mean introduces no meaningful leakage. Using the April mean instead
# would give virtually identical results given the magnitudes involved.

center_series <- function(x) x - mean(x, na.rm = TRUE)

evt_apr_model[, ret_mid_c := center_series(ret_mid)]
evt_apr_model[, ret_ia_c  := center_series(ret_ia)]

evt_may_model[, ret_mid_c := center_series(ret_mid)]
evt_may_model[, ret_ia_c  := center_series(ret_ia)]

evt_jun_model[, ret_mid_c := center_series(ret_mid)]
evt_jun_model[, ret_ia_c  := center_series(ret_ia)]

cat("Centred return means (should be ~0 by construction):\n")
cat(sprintf("  Apr ret_mid_c: %.2e  |  ret_ia_c: %.2e\n",
            mean(evt_apr_model$ret_mid_c), mean(evt_apr_model$ret_ia_c)))
cat(sprintf("  May ret_mid_c: %.2e  |  ret_ia_c: %.2e\n",
            mean(evt_may_model$ret_mid_c), mean(evt_may_model$ret_ia_c)))
cat(sprintf("  Jun ret_mid_c: %.2e  |  ret_ia_c: %.2e\n",
            mean(evt_jun_model$ret_mid_c), mean(evt_jun_model$ret_ia_c)))
```

---

## 5. Sanity Checks: Return Distributions

```{r return-distribution-plots}
# Visual comparison of raw vs IA centred returns for May and June.
# The IA returns should have a more stable variance than raw returns —
# intensity-adjustment removes activity-driven scale shifts.
# We thin by regular interval (not random) to preserve time structure.

thin_regular <- function(dt, N = 20000) {
  if (nrow(dt) <= N) return(dt)
  dt[round(seq(1, nrow(dt), length.out = N))]
}

plot_dt <- rbindlist(list(
  thin_regular(evt_may_model)[, .(timestamp, month = "May",
                                   lambda_hat, psi_hat, ret_mid_c, ret_ia_c)],
  thin_regular(evt_jun_model)[, .(timestamp, month = "Jun",
                                   lambda_hat, psi_hat, ret_mid_c, ret_ia_c)]
))

# Distribution comparison: raw vs IA
long_ret <- melt(plot_dt,
                 id.vars      = "month",
                 measure.vars = c("ret_mid_c", "ret_ia_c"),
                 variable.name = "series",
                 value.name    = "r")
long_ret[, series := fifelse(series == "ret_mid_c",
                             "Raw (centred)", "IA (centred)")]

ggplot(long_ret, aes(x = r)) +
  geom_histogram(bins = 120, fill = "steelblue", colour = NA, alpha = 0.8) +
  facet_grid(month ~ series, scales = "free_y") +
  labs(
    title    = "Event-level return distributions: Raw vs intensity-adjusted",
    subtitle = "IA returns scale raw returns by sqrt(psi_hat). Sample = 20k events per month.",
    x = "Return", y = "Count"
  ) +
  theme_minimal(base_size = 12)
```

```{r activity-volatility-correlation}
# Correlation check: does intensity-adjustment reduce the link between
# trading activity (lambda_hat) and absolute return magnitude?
# If IA works as intended, |ret_ia_c| should be less correlated with
# lambda_hat than |ret_mid_c| is.

corr_tab <- plot_dt[, .(
  corr_abs_raw_lambda = cor(abs(ret_mid_c), lambda_hat, use = "complete.obs"),
  corr_abs_ia_lambda  = cor(abs(ret_ia_c),  lambda_hat, use = "complete.obs")
), by = month]

cat("Correlation between |return| and intensity (lambda_hat):\n")
print(corr_tab)
cat("\nA reduction in correlation after IA confirms the adjustment\n")
cat("is removing activity-driven variance from the return series.\n")
```

---

## 6. 5-Minute Aggregation

```{r five-min-aggregation}
# Aggregate event-level returns to 5-minute buckets.
# Two tracks are produced for each month:
#
# (A) Volatility track: r_ia_5m_sum
#     Sum of event-level IA centred returns within each 5-min bucket.
#     Input to GARCH estimation in notebook 06.
#     Not directly invertible to raw returns — the scaling has been applied
#     at the event level before summation.
#
# (B) Risk track: r_ia_5m_inv
#     Raw 5-minute return × sqrt(mean psi_hat in bucket).
#     Invertible: r_raw_5m = r_ia_5m_inv / sqrt(psi_5m).
#     Used in VaR/ES backtesting (notebook 07) to map forecast quantiles
#     back to actual P&L in raw return units.
#
# psi_5m = mean of event-level psi_hat values within the bucket.
# Using mean (not median) because we want the quantity proportional to
# total expected duration in the bucket, not the central tendency.

make_5min_series <- function(dt, tz_use = "Europe/London") {
  dt <- data.table::copy(dt)
  setorder(dt, date, timestamp)

  # Assign each event to its 5-minute bucket (floor to nearest 300 seconds)
  dt[, t5 := as.POSIXct(
    floor(as.numeric(timestamp) / 300) * 300,
    origin = "1970-01-01", tz = tz_use
  )]

  out <- dt[, {
    list(
      r_raw_5m    = sum(ret_mid_c,  na.rm = TRUE),
      r_ia_5m_sum = sum(ret_ia_c,   na.rm = TRUE),
      psi_5m      = mean(psi_hat,   na.rm = TRUE),
      n_events    = .N
    )
  }, by = .(date, t5)][order(date, t5)]

  # Invertible IA return: raw 5-min return scaled by sqrt(mean psi in bucket)
  out[, scale_5m     := sqrt(pmax(psi_5m, 1e-12))]
  out[, r_ia_5m_inv  := r_raw_5m * scale_5m]

  out[]
}

r5_apr <- make_5min_series(evt_apr_model)
r5_may <- make_5min_series(evt_may_model)
r5_jun <- make_5min_series(evt_jun_model)

cat("5-minute series row counts:\n")
cat(sprintf("  Apr: %d buckets | May: %d buckets | Jun: %d buckets\n",
            nrow(r5_apr), nrow(r5_may), nrow(r5_jun)))

cat("\nEvents per 5-min bucket:\n")
cat("  Apr: "); cat(round(summary(r5_apr$n_events), 1)); cat("\n")
cat("  May: "); cat(round(summary(r5_may$n_events), 1)); cat("\n")
cat("  Jun: "); cat(round(summary(r5_jun$n_events), 1)); cat("\n")

cat("\nApril 5-min return summaries:\n")
cat("  Raw:       "); print(round(summary(r5_apr$r_raw_5m),    6))
cat("  IA sum:    "); print(round(summary(r5_apr$r_ia_5m_sum), 6))
cat("  IA inv:    "); print(round(summary(r5_apr$r_ia_5m_inv), 6))
```

---

## 7. 5-Minute Diagnostics

```{r events-per-bucket-plot}
# Distribution of event counts per 5-minute bucket.
# Heavy right tail expected — some buckets (US open, LSE close) have
# many more events than the median bucket. Confirms ACD intensity
# model is necessary — activity is highly non-uniform.

ggplot(r5_apr, aes(x = n_events)) +
  geom_histogram(bins = 60, fill = "steelblue", colour = NA, alpha = 0.85) +
  labs(
    title    = "April: event count per 5-minute bucket",
    subtitle = "Right-skewed distribution confirms non-uniform trading activity — motivates ACD modelling.",
    x = "Events in bucket", y = "Number of 5-minute buckets"
  ) +
  theme_minimal(base_size = 12)
```

```{r psi-5m-over-time}
# Bucket-level mean conditional duration (psi_5m) over time.
# Should show the intraday U-shape: longer durations at midday,
# shorter durations at open and close.
# Cross-month comparison shows May calmer (higher psi_5m) than April.

psi_long <- rbindlist(list(
  r5_apr[, .(t5, psi_5m, Month = "Apr")],
  r5_may[, .(t5, psi_5m, Month = "May")],
  r5_jun[, .(t5, psi_5m, Month = "Jun")]
))

ggplot(psi_long, aes(x = t5, y = psi_5m)) +
  geom_line(linewidth = 0.25, colour = "steelblue", alpha = 0.9) +
  facet_wrap(~Month, ncol = 1, scales = "free_x") +
  labs(
    title    = "Bucket-level conditional mean duration (ψ̂_5m) — April, May, June 2020",
    subtitle = "Higher ψ̂_5m = longer expected duration = lower intensity. May shows calmer market.",
    x = "Date", y = "Mean ψ̂ in 5-min bucket (seconds)"
  ) +
  theme_minimal(base_size = 12)
```

```{r activity-volatility-5m}
# At the 5-minute level: does intensity adjustment reduce the correlation
# between bucket activity (n_events, lambda_5m) and absolute returns?
# This is the key empirical test of whether the IA step is doing useful work.

mk_diag <- function(r5, label) {
  dt <- data.table::copy(r5)
  dt[, month     := label]
  dt[, lambda_5m := 1 / pmax(psi_5m, 1e-12)]
  dt[, abs_raw   := abs(r_raw_5m)]
  dt[, abs_iaSum := abs(r_ia_5m_sum)]
  dt[, abs_iaInv := abs(r_ia_5m_inv)]
  dt
}

dt5 <- rbindlist(list(
  mk_diag(r5_may, "May"),
  mk_diag(r5_jun, "Jun")
))

corr_5m <- dt5[, .(
  raw_vs_nEvents   = cor(abs_raw,   n_events,  use = "complete.obs"),
  iaSum_vs_nEvents = cor(abs_iaSum, n_events,  use = "complete.obs"),
  iaInv_vs_nEvents = cor(abs_iaInv, n_events,  use = "complete.obs"),
  raw_vs_lambda    = cor(abs_raw,   lambda_5m, use = "complete.obs"),
  iaSum_vs_lambda  = cor(abs_iaSum, lambda_5m, use = "complete.obs"),
  iaInv_vs_lambda  = cor(abs_iaInv, lambda_5m, use = "complete.obs")
), by = month]

cat("5-minute activity-volatility correlations:\n")
print(corr_5m)
cat("\nFor both IA tracks, reduced correlation with n_events and lambda_5m\n")
cat("confirms the intensity adjustment is removing activity-driven variance.\n")
```

```{r return-distribution-5m}
# Distribution comparison at 5-minute level: Raw vs IA sum vs IA inv.
# All three series go into different downstream uses — GARCH uses IA sum,
# VaR backtesting uses IA inv, raw is the benchmark.

dt_long <- melt(dt5,
                measure.vars  = c("r_raw_5m", "r_ia_5m_sum", "r_ia_5m_inv"),
                variable.name = "series",
                value.name    = "r")
dt_long[, series := fcase(
  series == "r_raw_5m",    "Raw",
  series == "r_ia_5m_sum", "IA sum (vol track)",
  series == "r_ia_5m_inv", "IA inv (risk track)"
)]

ggplot(dt_long, aes(x = r)) +
  geom_histogram(bins = 80, fill = "steelblue", colour = NA, alpha = 0.8) +
  facet_grid(month ~ series, scales = "free_y") +
  labs(
    title    = "5-minute return distributions: Raw vs IA sum vs IA inv",
    subtitle = "IA sum used for GARCH (notebook 06). IA inv used for VaR backtesting (notebook 07).",
    x = "5-minute return", y = "Count"
  ) +
  theme_minimal(base_size = 12)
```

---

## 8. Save Outputs

```{r save-outputs}
saveRDS(r5_apr, here::here("data", "processed", "r5_apr.rds"))
saveRDS(r5_may, here::here("data", "processed", "r5_may.rds"))
saveRDS(r5_jun, here::here("data", "processed", "r5_jun.rds"))

cat("Saved:\n")
cat(sprintf("  r5_apr.rds  — %d rows | cols: date, t5, r_raw_5m, r_ia_5m_sum,",
            nrow(r5_apr)))
cat(" r_ia_5m_inv, psi_5m, n_events\n")
cat(sprintf("  r5_may.rds  — %d rows\n", nrow(r5_may)))
cat(sprintf("  r5_jun.rds  — %d rows\n", nrow(r5_jun)))
cat("\nNotebook 06 reads r5_apr (GARCH estimation on r_ia_5m_sum)\n")
cat("and r5_may / r5_jun (OOS volatility forecasting).\n")
cat("Notebook 07 reads all three for VaR/ES backtesting on r_ia_5m_inv.\n")
```
