---
title: "Notebook 07 — VaR/ES Backtesting and Market Risk Evaluation"
author: "Sourodeep Roy"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r nb07_setup, message=FALSE, warning=FALSE}
library(data.table)
library(ggplot2)
library(rugarch)
library(sandwich)
library(lmtest)

# ── Anchor all paths to project root ─────────────────────────────────────────
proj_root <- rprojroot::find_rstudio_root_file()

# ── Load 5-min series from notebook 05 ───────────────────────────────────────
r5_apr <- readRDS(file.path(proj_root, "data/processed/r5_apr.rds"))
r5_may <- readRDS(file.path(proj_root, "data/processed/r5_may.rds"))
r5_jun <- readRDS(file.path(proj_root, "data/processed/r5_jun.rds"))

# ── Load GARCH fit objects and sigma series from notebook 06 ──────────────────
nb06 <- readRDS(file.path(proj_root, "data/processed/nb06_outputs.rds"))
list2env(nb06, envir = .GlobalEnv)

cat("Project root:", proj_root, "\n")
cat("Loaded from nb06:", paste(names(nb06), collapse = ", "), "\n\n")
cat("Rows: Apr =", nrow(r5_apr), "| May =", nrow(r5_may), "| Jun =", nrow(r5_jun), "\n")
```

```{r nb07_risk_series, message=FALSE, warning=FALSE}
# ── RISK TRACK SERIES ─────────────────────────────────────────────────────────
# Two tracks evaluated side by side, both assessed against RAW realised returns:
#
#   RAW model:  trained on r_raw_5m, VaR/ES on raw scale directly
#   IA model:   trained on r_ia_5m_inv, VaR/ES computed on IA scale,
#               then mapped back to raw scale via:
#                 VaR_raw = VaR_ia / scale_5m
#               where scale_5m = sqrt(psi_5m)  (exact invertible relationship)
#
# This design means both models are evaluated on identical realised losses,
# making the comparison fair and directly interpretable.

r_apr_raw_risk <- as.numeric(r5_apr$r_raw_5m)
r_apr_ia_risk  <- as.numeric(r5_apr$r_ia_5m_inv)

r_may_raw_risk <- as.numeric(r5_may$r_raw_5m)
r_may_ia_risk  <- as.numeric(r5_may$r_ia_5m_inv)

r_jun_raw_risk <- as.numeric(r5_jun$r_raw_5m)
r_jun_ia_risk  <- as.numeric(r5_jun$r_ia_5m_inv)

scale_may <- as.numeric(r5_may$scale_5m)
scale_jun  <- as.numeric(r5_jun$scale_5m)

stopifnot(
  all(is.finite(r_apr_raw_risk)), all(is.finite(r_apr_ia_risk)),
  all(is.finite(r_may_raw_risk)), all(is.finite(r_may_ia_risk)),
  all(is.finite(r_jun_raw_risk)), all(is.finite(r_jun_ia_risk)),
  all(is.finite(scale_may)), all(scale_may > 0),
  all(is.finite(scale_jun)),  all(scale_jun  > 0)
)

cat("Risk track lengths:\n")
cat("Apr:", length(r_apr_raw_risk),
    "| May:", length(r_may_raw_risk),
    "| Jun:", length(r_jun_raw_risk), "\n\n")

# ── Invertibility check ───────────────────────────────────────────────────────
# r_ia_inv = r_raw * scale  =>  r_raw = r_ia_inv / scale
# Correlation should be ~1 and MAE should be ~0.
r_may_raw_from_ia <- r_may_ia_risk / pmax(scale_may, 1e-12)
cat("Invertibility check (May):\n")
cat("  Corr(r_raw, r_raw_from_ia) =",
    round(cor(r_may_raw_risk, r_may_raw_from_ia, use = "complete.obs"), 6), "\n")
cat("  MAE =",
    signif(mean(abs(r_may_raw_risk - r_may_raw_from_ia), na.rm = TRUE), 4),
    "  [should be ~0]\n\n")

# ── Scale over time ───────────────────────────────────────────────────────────
scale_dt <- rbindlist(list(
  data.table(t5 = r5_may$t5, scale_5m = scale_may, Month = "May"),
  data.table(t5 = r5_jun$t5, scale_5m = scale_jun,  Month = "Jun")
))
ggplot(scale_dt, aes(x = t5, y = scale_5m)) +
  geom_line(linewidth = 0.25) +
  facet_wrap(~Month, ncol = 1, scales = "free_x") +
  labs(title = "Activity scaling: sqrt(psi_5m) over OOS months",
       x = "Time", y = "scale_5m = sqrt(psi_5m)") +
  theme_minimal()
```

```{r nb07_garch_fit_risk, message=FALSE, warning=FALSE}
# ── FIT GARCH ON RISK TRACK (April only) ─────────────────────────────────────
# We fit separately on the RISK track (r_ia_5m_inv), not the volatility track
# (r_ia_5m_sum) used in notebook 06. These are different series and can prefer
# different specs. RAW risk track uses Student-t (heavy tails in raw returns).
# IA risk track: test both Student-t and Normal since intensity adjustment
# compresses tails — BIC decides.

garch_specs <- list(
  sGARCH_std   = ugarchspec(
    variance.model     = list(model = "sGARCH",   garchOrder = c(1, 1)),
    mean.model         = list(armaOrder = c(0, 0), include.mean = FALSE),
    distribution.model = "std"),
  eGARCH_std   = ugarchspec(
    variance.model     = list(model = "eGARCH",   garchOrder = c(1, 1)),
    mean.model         = list(armaOrder = c(0, 0), include.mean = FALSE),
    distribution.model = "std"),
  gjrGARCH_std = ugarchspec(
    variance.model     = list(model = "gjrGARCH", garchOrder = c(1, 1)),
    mean.model         = list(armaOrder = c(0, 0), include.mean = FALSE),
    distribution.model = "std"),
  sGARCH_norm   = ugarchspec(
    variance.model     = list(model = "sGARCH",   garchOrder = c(1, 1)),
    mean.model         = list(armaOrder = c(0, 0), include.mean = FALSE),
    distribution.model = "norm"),
  eGARCH_norm   = ugarchspec(
    variance.model     = list(model = "eGARCH",   garchOrder = c(1, 1)),
    mean.model         = list(armaOrder = c(0, 0), include.mean = FALSE),
    distribution.model = "norm"),
  gjrGARCH_norm = ugarchspec(
    variance.model     = list(model = "gjrGARCH", garchOrder = c(1, 1)),
    mean.model         = list(armaOrder = c(0, 0), include.mean = FALSE),
    distribution.model = "norm")
)

fit_ic <- function(spec, data, label) {
  fit <- tryCatch(ugarchfit(spec, data = data, solver = "hybrid"), error = function(e) NULL)
  if (is.null(fit) || fit@fit$convergence != 0) return(NULL)
  ic <- infocriteria(fit)
  list(label = label, fit = fit, BIC = ic["Bayes", ])
}

# RAW: Student-t specs only (Normal is misspecified for raw returns)
res_raw <- lapply(names(garch_specs)[1:3], function(nm)
  fit_ic(garch_specs[[nm]], r_apr_raw_risk, nm))
res_raw <- Filter(Negate(is.null), res_raw)

# IA: all six specs (let BIC decide distribution)
res_ia <- lapply(names(garch_specs), function(nm)
  fit_ic(garch_specs[[nm]], r_apr_ia_risk, nm))
res_ia <- Filter(Negate(is.null), res_ia)

ic_raw <- rbindlist(lapply(res_raw, function(x)
  data.table(Track = "Raw", Model = x$label, BIC = round(x$BIC, 4))))
ic_ia  <- rbindlist(lapply(res_ia,  function(x)
  data.table(Track = "IA",  Model = x$label, BIC = round(x$BIC, 4))))

setorder(ic_raw, BIC); setorder(ic_ia, BIC)
cat("Risk track model selection (lower BIC preferred):\n")
print(rbindlist(list(ic_raw, ic_ia)))

# Select best
best_raw_risk <- res_raw[[which.min(sapply(res_raw, `[[`, "BIC"))]]
best_ia_risk  <- res_ia [[which.min(sapply(res_ia,  `[[`, "BIC"))]]

fit_raw_risk <- best_raw_risk$fit
fit_ia_risk  <- best_ia_risk$fit

cat("\nSelected: RAW =>", best_raw_risk$label, "| IA =>", best_ia_risk$label, "\n")
cat("\nRAW risk fit coefficients:\n"); print(round(coef(fit_raw_risk), 6))
cat("\nIA  risk fit coefficients:\n"); print(round(coef(fit_ia_risk),  6))
```

```{r nb07_garch_diagnostics, message=FALSE, warning=FALSE}
# ── GARCH RESIDUAL DIAGNOSTICS (April training fit) ───────────────────────────
# Standardised residuals z_t = r_t / sigma_t should be approximately iid.
# ACF(z) and ACF(z²): no remaining serial correlation.
# Ljung-Box on z and z²: formal test.
# QQ plot: distributional adequacy of the chosen innovation distribution.

diag_fit <- function(fit, label) {
  z <- as.numeric(na.omit(residuals(fit, standardize = TRUE)))

  lb_z  <- Box.test(z,   lag = 20, type = "Ljung-Box")
  lb_z2 <- Box.test(z^2, lag = 20, type = "Ljung-Box")

  cat("\n===", label, "===\n")
  cat("Ljung-Box on z  (lag 20): stat =", round(lb_z$statistic,  3),
      "| p =", round(lb_z$p.value,  4), "\n")
  cat("Ljung-Box on z² (lag 20): stat =", round(lb_z2$statistic, 3),
      "| p =", round(lb_z2$p.value, 4), "\n")

  op <- par(mfrow = c(1, 3))
  acf(z,   lag.max = 60, main = paste0(label, ": ACF(z)"))
  acf(z^2, lag.max = 60, main = paste0(label, ": ACF(z²)"))
  qqnorm(z, main = paste0(label, ": QQ of z")); qqline(z)
  par(op)

  invisible(z)
}

z_raw <- diag_fit(fit_raw_risk, "RAW risk model (April)")
z_ia  <- diag_fit(fit_ia_risk,  "IA risk model (April)")
```

```{r nb07_helpers, message=FALSE, warning=FALSE}
# ── HELPER FUNCTIONS ──────────────────────────────────────────────────────────
# Defined once here, used by all subsequent chunks.

# OOS sigma: freeze April parameters, filter through OOS month
oos_sigma_fixed <- function(fit_obj, r_oos) {
  spec_fx <- getspec(fit_obj)
  setfixed(spec_fx) <- as.list(coef(fit_obj))
  as.numeric(sigma(ugarchfilter(spec_fx, data = as.numeric(r_oos))))
}

# Kupiec (1995) unconditional coverage LR test
kupiec_uc <- function(I, alpha) {
  T <- length(I); x <- sum(I)
  eps <- 1e-12
  p   <- min(max(x / T, eps), 1 - eps)
  a   <- min(max(alpha,   eps), 1 - eps)
  LRuc <- -2 * ((T - x) * log(1 - a) + x * log(a) -
                (T - x) * log(1 - p) - x * log(p))
  list(LRuc = LRuc, p_uc = 1 - pchisq(LRuc, df = 1))
}

# Christoffersen (1998) independence test
christoffersen_ind <- function(I) {
  I <- as.integer(I)
  if (sum(I) == 0 || sum(I) == length(I))
    return(list(LRind = NA_real_, p_ind = NA_real_))

  n00 <- sum(I[-length(I)] == 0 & I[-1] == 0)
  n01 <- sum(I[-length(I)] == 0 & I[-1] == 1)
  n10 <- sum(I[-length(I)] == 1 & I[-1] == 0)
  n11 <- sum(I[-length(I)] == 1 & I[-1] == 1)

  eps <- 1e-12
  cl  <- function(p) min(max(p, eps), 1 - eps)
  pi01 <- cl(if ((n00 + n01) > 0) n01 / (n00 + n01) else 0)
  pi11 <- cl(if ((n10 + n11) > 0) n11 / (n10 + n11) else 0)
  pi   <- cl((n01 + n11) / (n00 + n01 + n10 + n11))

  ll0   <- (n00 + n10) * log(1 - pi)  + (n01 + n11) * log(pi)
  ll1   <- n00 * log(1 - pi01) + n01 * log(pi01) +
           n10 * log(1 - pi11) + n11 * log(pi11)
  LRind <- -2 * (ll0 - ll1)
  list(LRind = LRind, p_ind = 1 - pchisq(LRind, df = 1))
}

# Combined backtest row (Kupiec + Christoffersen CC)
var_backtest_row <- function(r, VaR, alpha, label) {
  I    <- as.integer(as.numeric(r) < as.numeric(VaR))
  T    <- length(I); hits <- sum(I)
  uc   <- kupiec_uc(I, alpha)
  ind  <- christoffersen_ind(I)
  LRcc <- if (!is.na(uc$LRuc) && !is.na(ind$LRind)) uc$LRuc + ind$LRind else NA_real_
  p_cc <- if (!is.na(LRcc)) 1 - pchisq(LRcc, df = 2) else NA_real_

  data.table(
    Sample   = label, alpha = alpha,
    T        = T,    hits  = hits, hit_rate = hits / T,
    exp_hits = round(alpha * T, 1),
    LRuc  = round(uc$LRuc,   3), p_uc  = round(uc$p_uc,  4),
    LRind = round(ind$LRind, 3), p_ind = round(ind$p_ind, 4),
    LRcc  = round(LRcc,      3), p_cc  = round(p_cc,      4)
  )
}

# FZ0 loss vector (Fissler & Ziegel 2016 — strictly consistent for (VaR, ES))
# Lower is better. Requires ES < 0 (left-tail convention).
fz0_loss_vec <- function(r, VaR, ES, alpha, eps = 1e-12) {
  r <- as.numeric(r); VaR <- as.numeric(VaR); ES <- as.numeric(ES)
  stopifnot(length(r) == length(VaR), length(r) == length(ES))
  ES_safe <- pmin(ES, -eps)
  I <- as.integer(r <= VaR)
  (1 / alpha) * (ES_safe - VaR) * (alpha - I) + log(-ES_safe)
}

# DM-HAC test on loss differential d_t
dm_hac <- function(d, L = NULL, alternative = c("greater", "two.sided")) {
  alternative <- match.arg(alternative)
  d <- as.numeric(na.omit(d)); T <- length(d)
  stopifnot(T > 30)
  if (is.null(L)) L <- floor(T^(1/3))
  fit <- lm(d ~ 1)
  V   <- NeweyWest(fit, lag = L, prewhite = FALSE, adjust = TRUE)
  ct  <- coeftest(fit, vcov. = V)
  tstat <- as.numeric(ct[1, 3])
  pval  <- if (alternative == "greater") 1 - pt(tstat, df = T - 1) else
                                         2 * (1 - pt(abs(tstat), df = T - 1))
  data.table(T = T, HAC_lag = L,
             mean_d  = round(as.numeric(ct[1, 1]), 6),
             t_stat  = round(tstat, 4),
             p_value = round(pval,  6))
}

# ES tail exceedance check
es_tail_check <- function(r, VaR, ES, label) {
  dt     <- data.table(r = as.numeric(r), VaR = as.numeric(VaR), ES = as.numeric(ES))
  dt_hit <- dt[is.finite(r) & is.finite(VaR) & is.finite(ES) & r < VaR]
  data.table(
    Sample         = label,
    breaches       = nrow(dt_hit),
    mean_r_breach  = if (nrow(dt_hit) > 0) round(mean(dt_hit$r),  6) else NA_real_,
    mean_ES_breach = if (nrow(dt_hit) > 0) round(mean(dt_hit$ES), 6) else NA_real_,
    # Ratio > 1: ES more negative than average realised loss (conservative)
    # Ratio < 1: ES less negative than average realised loss (dangerous)
    ES_cover_ratio = if (nrow(dt_hit) > 0)
      round(mean(dt_hit$ES) / mean(dt_hit$r), 4) else NA_real_
  )
}

cat("Helper functions loaded.\n")
```

```{r nb07_parametric_1pct, message=FALSE, warning=FALSE}
# ── PARAMETRIC VaR/ES AT ALPHA = 1% ──────────────────────────────────────────
# Closed-form multipliers from fitted innovation distributions.
#
# Student-t VaR multiplier: q_t(alpha, df=shape)
# Student-t ES multiplier:  E[Z | Z <= q] = -(f_t(q)/alpha) * (df + q²)/(df - 1)
#   This is exact and analytical — no simulation needed.
#
# Normal VaR multiplier: q_norm(alpha)
# Normal ES multiplier:  -phi(q_norm(alpha)) / alpha

alpha_1 <- 0.01

# Extract degrees of freedom from RAW fit
df_raw <- unname(coef(fit_raw_risk)["shape"])

# Check if IA fit used Student-t or Normal
ia_dist <- fit_ia_risk@model$modeldesc$distribution
cat("RAW innovation distribution:", fit_raw_risk@model$modeldesc$distribution,
    "| df =", round(df_raw, 4), "\n")
cat("IA  innovation distribution:", ia_dist, "\n\n")

# VaR multipliers
q_raw_1 <- qt(alpha_1, df = df_raw)
q_ia_1  <- if (ia_dist == "norm") qnorm(alpha_1) else
           qt(alpha_1, df = unname(coef(fit_ia_risk)["shape"]))

# Analytical ES multipliers (E[Z | Z <= q], negative value)
es_mult_t <- function(alpha, df) {
  q <- qt(alpha, df = df)
  -(dt(q, df = df) / alpha) * (df + q^2) / (df - 1)
}
es_mult_norm <- function(alpha) -dnorm(qnorm(alpha)) / alpha

es_raw_1 <- es_mult_t(alpha_1, df_raw)
es_ia_1  <- if (ia_dist == "norm") es_mult_norm(alpha_1) else
            es_mult_t(alpha_1, unname(coef(fit_ia_risk)["shape"]))

cat(sprintf("Parametric multipliers (alpha = %.2f):\n", alpha_1))
cat(sprintf("  RAW: q = %.4f | ES_mult = %.4f\n", q_raw_1, es_raw_1))
cat(sprintf("  IA:  q = %.4f | ES_mult = %.4f\n\n", q_ia_1,  es_ia_1))

# OOS sigma (April params frozen, filtered through May/Jun)
sig_may_raw <- oos_sigma_fixed(fit_raw_risk, r_may_raw_risk)
sig_jun_raw <- oos_sigma_fixed(fit_raw_risk, r_jun_raw_risk)
sig_may_ia  <- oos_sigma_fixed(fit_ia_risk,  r_may_ia_risk)
sig_jun_ia  <- oos_sigma_fixed(fit_ia_risk,  r_jun_ia_risk)

# VaR/ES on respective scales
VaR_may_raw_1 <- sig_may_raw * q_raw_1
VaR_jun_raw_1 <- sig_jun_raw * q_raw_1
ES_may_raw_1  <- sig_may_raw * es_raw_1
ES_jun_raw_1  <- sig_jun_raw * es_raw_1

VaR_may_ia_1_on_ia <- sig_may_ia * q_ia_1
VaR_jun_ia_1_on_ia <- sig_jun_ia * q_ia_1
ES_may_ia_1_on_ia  <- sig_may_ia * es_ia_1
ES_jun_ia_1_on_ia  <- sig_jun_ia * es_ia_1

# Map IA forecasts back to raw scale: VaR_raw = VaR_ia / scale_5m
VaR_may_ia_1 <- VaR_may_ia_1_on_ia / pmax(scale_may, 1e-12)
VaR_jun_ia_1 <- VaR_jun_ia_1_on_ia / pmax(scale_jun,  1e-12)
ES_may_ia_1  <- ES_may_ia_1_on_ia  / pmax(scale_may, 1e-12)
ES_jun_ia_1  <- ES_jun_ia_1_on_ia  / pmax(scale_jun,  1e-12)

stopifnot(all(VaR_may_raw_1 < 0), all(VaR_may_ia_1 < 0),
          all(VaR_jun_raw_1 < 0), all(VaR_jun_ia_1 < 0))
cat("Parametric 1% VaR/ES vectors built. All VaR values negative. \n")
```

```{r nb07_backtest_1pct, message=FALSE, warning=FALSE}
# ── BACKTESTS AT ALPHA = 1% ───────────────────────────────────────────────────
# Kupiec (1995): H0 = hit rate equals alpha (unconditional coverage).
# Christoffersen (1998): additionally tests for independence of hits.
# CC test = UC + IND (joint, df=2).
# Both models evaluated on the SAME realised raw returns.

r_all_raw   <- c(r_may_raw_risk, r_jun_raw_risk)
VaR_all_raw_1 <- c(VaR_may_raw_1, VaR_jun_raw_1)
VaR_all_ia_1  <- c(VaR_may_ia_1,  VaR_jun_ia_1)
ES_all_raw_1  <- c(ES_may_raw_1,  ES_jun_raw_1)
ES_all_ia_1   <- c(ES_may_ia_1,   ES_jun_ia_1)

bt_1pct <- rbindlist(list(
  var_backtest_row(r_may_raw_risk, VaR_may_raw_1, alpha_1, "May RAW (param 1%)"),
  var_backtest_row(r_may_raw_risk, VaR_may_ia_1,  alpha_1, "May IA-mapped (param 1%)"),
  var_backtest_row(r_jun_raw_risk, VaR_jun_raw_1, alpha_1, "Jun RAW (param 1%)"),
  var_backtest_row(r_jun_raw_risk, VaR_jun_ia_1,  alpha_1, "Jun IA-mapped (param 1%)"),
  var_backtest_row(r_all_raw, VaR_all_raw_1, alpha_1, "May+Jun RAW (param 1%)"),
  var_backtest_row(r_all_raw, VaR_all_ia_1,  alpha_1, "May+Jun IA-mapped (param 1%)")
))

cat("VaR backtests at alpha = 1%:\n")
print(bt_1pct)

# ── Plot: May returns + both VaR lines + breach dots ─────────────────────────
dtp <- data.table(
  t       = seq_along(r_may_raw_risk),
  r       = r_may_raw_risk,
  VaR_raw = VaR_may_raw_1,
  VaR_ia  = VaR_may_ia_1
)
dtp[, hit_raw := r < VaR_raw]

ggplot(dtp, aes(x = t)) +
  geom_line(aes(y = r),       colour = "grey50",    linewidth = 0.3) +
  geom_line(aes(y = VaR_raw), colour = "tomato",    linewidth = 0.55) +
  geom_line(aes(y = VaR_ia),  colour = "steelblue", linewidth = 0.55,
            linetype = "dashed") +
  geom_point(data = dtp[hit_raw == TRUE], aes(y = r),
             colour = "tomato", size = 1.2) +
  labs(
    title    = "May (1% VaR): realised raw returns with VaR forecasts",
    subtitle = "Red solid = RAW | Blue dashed = IA mapped to raw scale | Dots = RAW breaches",
    x        = "5-min bucket index",
    y        = "Return (raw scale)"
  ) +
  theme_minimal()
```

```{r nb07_fhs_5pct, message=FALSE, warning=FALSE}
# ── FILTERED HISTORICAL SIMULATION (FHS) AT ALPHA = 5% ───────────────────────
# McNeil & Frey (2000): standardise returns by GARCH sigma, use the empirical
# quantile of April standardised residuals as the VaR multiplier.
# Advantage: no distributional assumption on the innovation tail shape.
# ES multiplier: E[z | z <= q_fhs] computed from the same April residuals.
#
# FHS VaR_t = sigma_t * q_fhs
# FHS ES_t  = sigma_t * mean(z_april | z_april <= q_fhs)

alpha_5 <- 0.05

# April standardised residuals (in-sample from frozen-param fits)
z_apr_raw <- as.numeric(na.omit(residuals(fit_raw_risk, standardize = TRUE)))
z_apr_ia  <- as.numeric(na.omit(residuals(fit_ia_risk,  standardize = TRUE)))

# Empirical quantile at alpha (FHS VaR multiplier)
q_raw_fhs <- as.numeric(quantile(z_apr_raw, alpha_5))
q_ia_fhs  <- as.numeric(quantile(z_apr_ia,  alpha_5))

# Empirical ES multiplier: conditional mean below the quantile
es_raw_fhs <- mean(z_apr_raw[z_apr_raw <= q_raw_fhs])
es_ia_fhs  <- mean(z_apr_ia [z_apr_ia  <= q_ia_fhs ])

cat(sprintf("FHS multipliers (alpha = %.2f):\n", alpha_5))
cat(sprintf("  RAW: q_fhs = %.4f | E[z|z<=q] = %.4f | ES/VaR = %.4f\n",
            q_raw_fhs, es_raw_fhs, es_raw_fhs / q_raw_fhs))
cat(sprintf("  IA:  q_fhs = %.4f | E[z|z<=q] = %.4f | ES/VaR = %.4f\n\n",
            q_ia_fhs, es_ia_fhs, es_ia_fhs / q_ia_fhs))

# FHS VaR/ES on each track's scale
VaR_may_raw_5 <- sig_may_raw * q_raw_fhs
VaR_jun_raw_5 <- sig_jun_raw * q_raw_fhs
ES_may_raw_5  <- sig_may_raw * es_raw_fhs
ES_jun_raw_5  <- sig_jun_raw * es_raw_fhs

VaR_may_ia_5_on_ia <- sig_may_ia * q_ia_fhs
VaR_jun_ia_5_on_ia <- sig_jun_ia * q_ia_fhs
ES_may_ia_5_on_ia  <- sig_may_ia * es_ia_fhs
ES_jun_ia_5_on_ia  <- sig_jun_ia * es_ia_fhs

# Map IA back to raw scale
VaR_may_ia_5 <- VaR_may_ia_5_on_ia / pmax(scale_may, 1e-12)
VaR_jun_ia_5 <- VaR_jun_ia_5_on_ia / pmax(scale_jun,  1e-12)
ES_may_ia_5  <- ES_may_ia_5_on_ia  / pmax(scale_may, 1e-12)
ES_jun_ia_5  <- ES_jun_ia_5_on_ia  / pmax(scale_jun,  1e-12)

stopifnot(all(VaR_may_raw_5 < 0), all(VaR_may_ia_5 < 0))
cat("FHS 5% VaR/ES vectors built. All VaR values negative. \n")
```

```{r nb07_backtest_5pct, message=FALSE, warning=FALSE}
# ── BACKTESTS AT ALPHA = 5% (FHS) ────────────────────────────────────────────

r_all_raw     <- c(r_may_raw_risk, r_jun_raw_risk)
VaR_all_raw_5 <- c(VaR_may_raw_5, VaR_jun_raw_5)
VaR_all_ia_5  <- c(VaR_may_ia_5,  VaR_jun_ia_5)
ES_all_raw_5  <- c(ES_may_raw_5,  ES_jun_raw_5)
ES_all_ia_5   <- c(ES_may_ia_5,   ES_jun_ia_5)

bt_5pct <- rbindlist(list(
  var_backtest_row(r_may_raw_risk, VaR_may_raw_5, alpha_5, "May RAW (FHS 5%)"),
  var_backtest_row(r_may_raw_risk, VaR_may_ia_5,  alpha_5, "May IA-mapped (FHS 5%)"),
  var_backtest_row(r_jun_raw_risk, VaR_jun_raw_5, alpha_5, "Jun RAW (FHS 5%)"),
  var_backtest_row(r_jun_raw_risk, VaR_jun_ia_5,  alpha_5, "Jun IA-mapped (FHS 5%)"),
  var_backtest_row(r_all_raw, VaR_all_raw_5, alpha_5, "May+Jun RAW (FHS 5%)"),
  var_backtest_row(r_all_raw, VaR_all_ia_5,  alpha_5, "May+Jun IA-mapped (FHS 5%)")
))

cat("VaR backtests at alpha = 5% (FHS):\n")
print(bt_5pct)
```

```{r nb07_cp_ci_plot, message=FALSE, warning=FALSE}
# ── CLOPPER–PEARSON CONFIDENCE INTERVALS ON HIT RATES ────────────────────────
# Standard in regulatory VaR validation (BCBS 2019).
# 95% CI for the binomial hit proportion — tells us whether observed coverage
# is statistically consistent with the target alpha.
# If the CI contains alpha: model is well-calibrated.
# If the CI is above alpha: model under-covers (too few breaches, conservative).
# If the CI is below alpha: model over-covers (too many breaches, dangerous).

binom_ci_row <- function(row) {
  x <- row$hits; n <- row$T; a <- row$alpha
  eps <- 1 - 0.95
  lo <- if (x == 0) 0 else qbeta(eps / 2,       x,     n - x + 1)
  hi <- if (x == n) 1 else qbeta(1 - eps / 2, x + 1,   n - x)
  data.table(
    Sample       = row$Sample,
    alpha_target = a,
    T            = n,
    hits         = x,
    hit_rate     = round(x / n, 5),
    ci_lo        = round(lo, 5),
    ci_hi        = round(hi, 5)
  )
}

all_bt <- rbindlist(list(bt_1pct, bt_5pct), use.names = TRUE, fill = TRUE)
tab_ci <- rbindlist(lapply(1:nrow(all_bt), function(i) binom_ci_row(all_bt[i])))

cat("Hit-rate calibration with 95% Clopper-Pearson CI:\n")
print(tab_ci)

ggplot(tab_ci, aes(x = hit_rate, y = reorder(Sample, -alpha_target))) +
  geom_point(aes(colour = factor(alpha_target)), size = 2.5) +
  geom_errorbarh(aes(xmin = ci_lo, xmax = ci_hi,
                     colour = factor(alpha_target)), height = 0.22) +
  geom_vline(xintercept = 0.01, linetype = "dashed",
             colour = "tomato", alpha = 0.7) +
  geom_vline(xintercept = 0.05, linetype = "dashed",
             colour = "steelblue", alpha = 0.7) +
  scale_colour_manual(values = c("0.01" = "tomato", "0.05" = "steelblue"),
                      labels = c("0.01" = "α=1%", "0.05" = "α=5%"),
                      name   = "Target α") +
  labs(title = "VaR hit-rate calibration with 95% Clopper–Pearson CI",
       x = "Observed hit-rate (with 95% CI)", y = NULL) +
  theme_minimal() + theme(legend.position = "bottom")
```

```{r nb07_basel_traffic_light, message=FALSE, warning=FALSE}
# ── BASEL II/III TRAFFIC LIGHT ────────────────────────────────────────────────
# For a standard 250-day 99% VaR backtest:
#   Green  : 0–4  breaches — model accepted
#   Yellow : 5–9  breaches — each zone adds a "plus factor" to capital multiplier
#   Red    : 10+  breaches — model automatically rejected
#
# We adapt this to 5-minute data by computing the annualised equivalent breach
# count: annual_equiv = hits × (250 × 102) / T
# where 102 = 5-min buckets per trading day (8.5 hours × 12 buckets/hour).
# This contextualises results within Basel regulatory expectations.

buckets_per_day <- 102L
trading_days    <- 250L

bt_combined_1pct <- bt_1pct[grepl("May\\+Jun", Sample)]
bt_combined_1pct[, annual_equiv := round(hits * (trading_days * buckets_per_day) / T, 1)]
bt_combined_1pct[, Basel_zone   := fcase(
  annual_equiv <= 4,  "Green",
  annual_equiv <= 9,  "Yellow",
  default = "Red"
)]

cat("Basel traffic light (annualised equivalent breach count at 1%):\n")
print(bt_combined_1pct[, .(Sample, T, hits, hit_rate, annual_equiv, Basel_zone)])
```

```{r nb07_es_diagnostics, message=FALSE, warning=FALSE}
# ── ES TAIL DIAGNOSTICS ───────────────────────────────────────────────────────
# Conditional on VaR breaches, the ES should approximate the mean realised
# tail loss. ES_cover_ratio = mean(ES | breach) / mean(r | breach).
# Ideal: ratio ≈ 1. Ratio > 1: ES overstates tail severity (conservative).
# Ratio < 1: ES understates tail severity (dangerous).
# This is standard in internal model validation under FRTB (BCBS 2016).

es_tab_1 <- rbindlist(list(
  es_tail_check(r_may_raw_risk, VaR_may_raw_1, ES_may_raw_1, "May RAW (param 1%)"),
  es_tail_check(r_may_raw_risk, VaR_may_ia_1,  ES_may_ia_1,  "May IA-mapped (param 1%)"),
  es_tail_check(r_jun_raw_risk, VaR_jun_raw_1, ES_jun_raw_1, "Jun RAW (param 1%)"),
  es_tail_check(r_jun_raw_risk, VaR_jun_ia_1,  ES_jun_ia_1,  "Jun IA-mapped (param 1%)")
))

es_tab_5 <- rbindlist(list(
  es_tail_check(r_may_raw_risk, VaR_may_raw_5, ES_may_raw_5, "May RAW (FHS 5%)"),
  es_tail_check(r_may_raw_risk, VaR_may_ia_5,  ES_may_ia_5,  "May IA-mapped (FHS 5%)"),
  es_tail_check(r_jun_raw_risk, VaR_jun_raw_5, ES_jun_raw_5, "Jun RAW (FHS 5%)"),
  es_tail_check(r_jun_raw_risk, VaR_jun_ia_5,  ES_jun_ia_5,  "Jun IA-mapped (FHS 5%)")
))

cat("ES tail diagnostics — 1% parametric:\n"); print(es_tab_1)
cat("\nES tail diagnostics — 5% FHS:\n");       print(es_tab_5)
cat("\n  ES_cover_ratio close to 1 = well-calibrated ES\n")
cat("  Ratio > 1 = conservative (ES more negative than average realised loss)\n")
cat("  Ratio < 1 = dangerous (ES understates tail)\n")
```

```{r nb07_fz0_dm, message=FALSE, warning=FALSE}
# ── FZ0 JOINT VaR–ES LOSS AND DM-HAC TEST ────────────────────────────────────
# Fissler & Ziegel (2016): FZ0 is strictly consistent for (VaR, ES) jointly.
# It is the preferred evaluation criterion under FRTB for internal ES models.
# We test H1: IA-mapped model has lower (better) FZ0 loss than Raw.
# d_t = FZ0_raw(t) - FZ0_ia(t) ; positive mean => IA wins.

# 1% parametric loss series
L_raw_may_1 <- fz0_loss_vec(r_may_raw_risk, VaR_may_raw_1, ES_may_raw_1, alpha_1)
L_ia_may_1  <- fz0_loss_vec(r_may_raw_risk, VaR_may_ia_1,  ES_may_ia_1,  alpha_1)
L_raw_jun_1 <- fz0_loss_vec(r_jun_raw_risk, VaR_jun_raw_1, ES_jun_raw_1, alpha_1)
L_ia_jun_1  <- fz0_loss_vec(r_jun_raw_risk, VaR_jun_ia_1,  ES_jun_ia_1,  alpha_1)

d_may_1 <- L_raw_may_1 - L_ia_may_1
d_jun_1 <- L_raw_jun_1 - L_ia_jun_1
d_all_1 <- c(d_may_1, d_jun_1)

# 5% FHS loss series
L_raw_may_5 <- fz0_loss_vec(r_may_raw_risk, VaR_may_raw_5, ES_may_raw_5, alpha_5)
L_ia_may_5  <- fz0_loss_vec(r_may_raw_risk, VaR_may_ia_5,  ES_may_ia_5,  alpha_5)
L_raw_jun_5 <- fz0_loss_vec(r_jun_raw_risk, VaR_jun_raw_5, ES_jun_raw_5, alpha_5)
L_ia_jun_5  <- fz0_loss_vec(r_jun_raw_risk, VaR_jun_ia_5,  ES_jun_ia_5,  alpha_5)

d_may_5 <- L_raw_may_5 - L_ia_may_5
d_jun_5 <- L_raw_jun_5 - L_ia_jun_5
d_all_5 <- c(d_may_5, d_jun_5)

# Mean FZ0 table
fz0_tab <- rbindlist(list(
  data.table(alpha=alpha_1, Sample="May RAW (param 1%)",          FZ0=mean(L_raw_may_1)),
  data.table(alpha=alpha_1, Sample="May IA-mapped (param 1%)",    FZ0=mean(L_ia_may_1)),
  data.table(alpha=alpha_1, Sample="Jun RAW (param 1%)",          FZ0=mean(L_raw_jun_1)),
  data.table(alpha=alpha_1, Sample="Jun IA-mapped (param 1%)",    FZ0=mean(L_ia_jun_1)),
  data.table(alpha=alpha_1, Sample="May+Jun RAW (param 1%)",      FZ0=mean(c(L_raw_may_1, L_raw_jun_1))),
  data.table(alpha=alpha_1, Sample="May+Jun IA-mapped (param 1%)",FZ0=mean(c(L_ia_may_1,  L_ia_jun_1))),
  data.table(alpha=alpha_5, Sample="May RAW (FHS 5%)",            FZ0=mean(L_raw_may_5)),
  data.table(alpha=alpha_5, Sample="May IA-mapped (FHS 5%)",      FZ0=mean(L_ia_may_5)),
  data.table(alpha=alpha_5, Sample="Jun RAW (FHS 5%)",            FZ0=mean(L_raw_jun_5)),
  data.table(alpha=alpha_5, Sample="Jun IA-mapped (FHS 5%)",      FZ0=mean(L_ia_jun_5)),
  data.table(alpha=alpha_5, Sample="May+Jun RAW (FHS 5%)",        FZ0=mean(c(L_raw_may_5, L_raw_jun_5))),
  data.table(alpha=alpha_5, Sample="May+Jun IA-mapped (FHS 5%)",  FZ0=mean(c(L_ia_may_5,  L_ia_jun_5)))
))
cat("FZ0 joint VaR-ES loss (lower is better):\n"); print(fz0_tab)

# DM-HAC tests
dm_fz <- rbindlist(list(
  cbind(alpha=alpha_1, Sample="May",     dm_hac(d_may_1, alternative="greater")),
  cbind(alpha=alpha_1, Sample="Jun",     dm_hac(d_jun_1, alternative="greater")),
  cbind(alpha=alpha_1, Sample="May+Jun", dm_hac(d_all_1, alternative="greater")),
  cbind(alpha=alpha_5, Sample="May",     dm_hac(d_may_5, alternative="greater")),
  cbind(alpha=alpha_5, Sample="Jun",     dm_hac(d_jun_5, alternative="greater")),
  cbind(alpha=alpha_5, Sample="May+Jun", dm_hac(d_all_5, alternative="greater"))
))
cat("\nDM-HAC on FZ0: d_t = FZ0_raw - FZ0_IA (H1: IA better => mean(d)>0):\n")
print(dm_fz)

# Win rates
win_fz <- rbindlist(lapply(
  list(list("May 1%",     d_may_1),
       list("Jun 1%",     d_jun_1),
       list("May+Jun 1%", d_all_1),
       list("May 5%",     d_may_5),
       list("Jun 5%",     d_jun_5),
       list("May+Jun 5%", d_all_5)),
  function(x) data.table(
    Sample     = x[[1]],
    T          = length(x[[2]]),
    WinRate_IA = round(mean(x[[2]] > 0), 4),
    TieRate    = round(mean(x[[2]] == 0), 4))
))
cat("\nDirectional win-rate (FZ0):\n"); print(win_fz)
```

```{r nb07_cumulative_fz0, message=FALSE, warning=FALSE}
# ── CUMULATIVE FZ0 ADVANTAGE PLOT ─────────────────────────────────────────────
# Persistent upward slope = IA-mapped model consistently better across time.
# Sharp rises on volatile days show ACD-GARCH provides the most benefit
# precisely when accurate risk measurement matters most.
# A flat section means both models are equally competitive in quiet periods.

dt_cum <- rbindlist(list(
  data.table(t      = seq_along(d_all_1),
             cum_adv = cumsum(d_all_1),
             alpha  = "1% parametric"),
  data.table(t      = seq_along(d_all_5),
             cum_adv = cumsum(d_all_5),
             alpha  = "5% FHS")
))

ggplot(dt_cum, aes(x = t, y = cum_adv)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  facet_wrap(~alpha, ncol = 1, scales = "free_y") +
  labs(
    title    = "Cumulative FZ0 advantage: Raw minus IA-mapped (May+Jun)",
    subtitle = "Upward trend => IA-mapped model has lower (better) joint VaR-ES loss over time",
    x        = "5-min bucket index (May+Jun combined)",
    y        = "Cumulative sum of d_t"
  ) +
  theme_minimal()
```

```{r nb07_summary_table, message=FALSE, warning=FALSE}
# ── COMBINED SUMMARY TABLE ────────────────────────────────────────────────────
# Dissertation Results section table — consolidates all VaR backtest results.

all_bt <- rbindlist(list(bt_1pct, bt_5pct), use.names = TRUE, fill = TRUE)
all_bt[, Method := fifelse(grepl("FHS",   Sample), "FHS", "Parametric")]
all_bt[, Month  := fcase(
  grepl("May\\+Jun", Sample), "May+Jun",
  grepl("^May",      Sample), "May",
  grepl("^Jun",      Sample), "Jun"
)]
all_bt[, Model  := fifelse(grepl("IA", Sample), "IA-mapped", "RAW")]
setorder(all_bt, alpha, Month, Method, Model)

cat("\n=== COMBINED VaR BACKTEST SUMMARY ===\n")
print(all_bt[, .(alpha, Method, Month, Model,
                 T, hits, hit_rate,
                 exp_hits, LRuc, p_uc, LRcc, p_cc)])
```

```{r nb07_save, message=FALSE, warning=FALSE}
# ── SAVE OUTPUTS ──────────────────────────────────────────────────────────────
saveRDS(list(
  bt_1pct     = bt_1pct,
  bt_5pct     = bt_5pct,
  fz0_tab     = fz0_tab,
  dm_fz       = dm_fz,
  win_fz      = win_fz,
  es_tab_1    = es_tab_1,
  es_tab_5    = es_tab_5,
  all_bt      = all_bt
), file.path(proj_root, "data/processed/nb07_outputs.rds"))

cat("Saved: data/processed/nb07_outputs.rds\n")
```
## Conclusion

This notebook completes the out-of-sample risk evaluation of the ACD-GARCH framework
introduced in notebooks 01–06. The key findings are summarised below.

### VaR Backtesting

Both the RAW and IA-mapped models exhibit conservative coverage under the frozen-parameter
OOS design, reflecting the higher volatility regime of the April training window relative
to the May–June test period. At 1%, Kupiec's LR test rejects for all specifications
(hit rates of 0.2–0.3% against a 1% target). At 5% using Filtered Historical Simulation,
the Clopper–Pearson 95% confidence intervals contain the target coverage level in all cases,
indicating that under-coverage at this level is not statistically distinguishable from correct
calibration. Hit independence (Christoffersen) passes throughout at both alpha levels,
confirming that the GARCH variance dynamics are correctly specified even where the
calibration level is off.

### ES Diagnostics

Parametric ES at 1% understates conditional tail severity (ES cover ratios of 0.78–0.93),
a consequence of the over-conservative VaR threshold selecting a non-representative breach
sample. FHS ES at 5% achieves cover ratios of 1.01–1.06 across both months and both models —
accurate conditional tail estimation with a small conservative margin.

### FZ0 Joint VaR–ES Loss

The IA-mapped model produces lower FZ0 joint loss than the RAW model in all 12 evaluation
configurations (both alpha levels, both months, combined sample). DM-HAC t-statistics range
from 9.5 to 36.4 (p < 0.001 in all cases) after Newey-West HAC correction with T^(1/3) lag
selection. The IA-mapped model achieves lower point-in-time FZ0 loss in 90–97% of 5-minute
intervals at the 1% level and 69–84% at the 5% level. The cumulative FZ0 advantage plot
shows a monotonically increasing trajectory with no sustained reversals across the full
4,178-bucket OOS window.

### Overall Assessment

The ACD intensity adjustment produces measurable and statistically significant improvements
in joint VaR–ES forecast quality at both the 1% and 5% significance levels. The improvement
is pervasive across time rather than concentrated in specific volatile periods, and is robust
to the choice of alpha level and estimation method (parametric vs FHS). The primary limitation
of the current evaluation is the single-month training window: models calibrated on April
carry over-estimated variance into a calmer OOS period, producing conservative VaR thresholds
at both alpha levels. A rolling or expanding window design would be expected to improve
absolute calibration while preserving the relative IA advantage documented here.